---
title: "Foliar fungal symbionts in sympatric yellow monkeyflowers along elevation gradients in the Sierra Nevada"
date: last-modified
date-format: "MMMM D, YYYY"
bibliography: references.yaml
#csl: apa.csl
csl: molecular-ecology.csl

author:
  - name: Bolívar Aponte Rolón
    orcid: 0000-0002-2544-4551
    email: bolivar@jhmi.edu
    corresponding: true
    affiliations:
      # - ref: jh
      - ref: tu

  - name: Kathleen G. Ferris
    email: kferris@tulane.edu
    corresponding: false
    affiliations:
      - ref: tu

  - name: Sunshine A. Van Bael
    orcid: 0000-0001-7317-3533
    email: svanbael@tulane.edu
    corresponding: false
    affiliation:
      - ref: tu

affiliations:
  # - id: jh
  #   name: Johns Hopkins All Childrens's Hospital, 600 6th Street South, St. Petersburg, FL 33701
  - id: tu
    name: Department of Ecology and Evolutionary Biology, Tulane University, 6823 St. Charles Avenue, New Orleans, LA 70118


filters:
  - authors-block

format:
  docx:
    reference-doc: custom-reference-doc.docx
    number-sections: true
    tbl-colwidths: true
    citation-package: biblatex
  # pdf:
  #   template:  WileyDesign/Optimal-Design-layout/Optimal-Design-layout.tex
    
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Table S
      space-before-numbering: false
      caption-location: top
      latex-list-of-description: Supplementary Table

editor: source
execute:
  echo: false
warning: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true
#| tidy: true

source("R/utils/setup.R")

path <- here::here()
data <- load_datasets(path) |> prepare_derived_data()
list2env(data, envir = .GlobalEnv)

# Comparison lists are already available:
# species_comparisons, elevation_comparisons

# Traits for dbRDA
dbrda_hell_matrix <- rrfy_hell_matrix |>
  column_to_rownames(var = "X") |>
  select(-c(1, 233:256))

dbrda_traits <- rrfy_hell_matrix |>
  select(
    X,
    Unique_ID,
    Site,
    Species,
    Elevation_m,
    Elevation_cat,
    Sample_Date,
    logACI,
    logLT,
    logLPS,
    logLMA,
    logLBI
  ) |>
  unite(Species_Elevation, Species, Elevation_cat, sep = "_", remove = FALSE) # To model interactions visually.

# Dataset with missing values imputed
dbrda_traits_imp <- dbrda_traits |>
  mutate(across(c(, 8:13), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
```

## Keywords

# Abstract

Microbial symbionts have the potential to contribute to host-plant's ecological and evolutionary success, especially in plants' adaptions to harsh environments, however their role has often been overlooked. Conversely, how host local adaptation (e.g., trait divergence across elevation) shapes the composition of associated microbial symbiont communities remains poorly understood. We explored how foliar endophytic fungi (FEF) abundance, richness, and community composition in three sympatric Monkeyflowers, an ecologically diverse group of flowering plants, change across elevation in the Sierra Nevada, CA, USA. We asked: Q1) Are there differences in leaf functional traits among sympatric *Mimulus* species populations at similar elevations? Q2) Within each host species, how do leaf traits vary along the elevation gradient? Q3) Do host species and elevation jointly shape FEF richness and community composition, and do leaf functional traits predict FEF β-diversity? Q4) Within each host species, does FEF community composition change with geographic distance? We collected *M. guttatus*, *M. nasutus*, and *M.laciniatus* individuals from natural populations across the Sierra Nevada, measured leaf functional traits and used ITS sequencing to describe the leaf endophyte community. We found significant associations of FEF community composition with host species, and elevation, suggesting that these factors influence fungal community composition. Furthermore, FEF community dissimilarity was correlated with geographic distance indicating isolation by distance and limited dispersal of fungal endophytes. We detected the prevalence of *Vishniacozyma victoriae*, an endophyte found in Antarctica, across all Sierran *Mimulus* populations. The presence of *V. victoriae* could play a role in the adaptation of *Mimulus* to cold, high elevation environments.  Our findings offer novel insights into the intricate interactions between fungal endophyte communities, plant traits, and elevation.


# Introduction

Microbial dispersal is challenging to trace. The general assumption is that most, if not all, microbes are ubiquitous. Nevertheless, the dispersal of microbial communities is not random as it is the result of neutral and selective processes influenced by environmental conditions, geographic distance, and host species [@rebolledagomez2019]. The co-evolution and symbiotic interactions of plant microbes, such as fungi [@remy1994; @wang2006; @field2015; reviewed in @peay2016] and bacteria [@soltis1995; @adams2002; @delaux2015], aided plants’ colonization of land across multiple varying climates. However, little is still known about how plant microbiomes influence or are influenced by local adaptation of their host species. Geographic location and environmental conditions are known to influence community composition of symbiotic fungi, with varying effects across different fungal guilds [@bowman2021; @kivlin2022]. Previous studies have suggested that foliar endophytic fungi (FEF), which live inside plant leaves, can alter plant traits under stressful conditions such as drought [@song2016; @giauque2019; @giauque2013] and therefore may contribute to local adaptation. Given these close associations between host and symbiont the resulting FEF community is influenced by the interplay of environmental conditions, geographic distance and host species.
Host traits and host identity can influence FEF communities in a variety of ways. Leaf functional traits, such as leaf thickness, leaf mass per area, and leaf toughness are known to influence plant's response to abiotic factors like light, temperature, and hydraulic constraints [@nicotra2011; @oguchi2018]. These traits serve as host-imposed filters that play a role in structuring FEF communities [@saunders2010; @tellez2022]. Jointly, host identity has been shown to structure the foliar microbiomes of plants. For example, a study of the classical plant model, *Arabidopsis thaliana*, revealed that bacterial and fungal communities in leaves were shaped by host genotype [@horton2014]. Recently, Karasov et al. [-@karasov2024], found further support of this hypothesis, showing that the bacterial leaf microbiome in diverse European populations of *A. thaliana* is influenced by genetic variation in the host. At the local scale, *A. thaliana* genotype can impose strong selection on the composition of its leaf microbiome, but at the continental scale, while genotype still plays a role, geography and abiotic factors also contribute strongly to symbiont community composition [@karasov2024]. The patterns observed in this model system encourage testing these hypotheses in other plant species to determine their broader applicability. The *Mimulus guttatus* species complex, a closely related interfertile group of species with considerable morphological and ecological variation, is an ideal model system to explore questions about FEF communities and their relationship with host speecies, functional traits, and environmental variation [@vickery1978; @wu2008; with bacterial endophityc communities @salinasaguilar2022]. 

Previous studies have examined floral and root microbial symbionts in *Mimulus* species, yet little is known about their foliar symbionts [@belisle2012; @rebolledagomez2019; @mcintosh2024]. Aboveground, Beslile et al. [-@belisle2012] reported on the distribution of diverse fungal communities in the flower nectar of *Mimulus aurantiacus*. The authors considered flowers as islands in a metapopulation system and found that the frequency of micro-fungi (i.e., yeasts) was significantly correlated to the location of the plant and marginally correlated with the density of the flowers in the plant [@belisle2012]. Another study by Rebolleda et al [-@rebolledagomez2019], examined epiphytic bacterial communities of different flower parts in *Mimulus guttatus* and found that pollinator choice drives strong micro-environmental selection of bacterial communities, overwhelming the effects of geographic distance. Belowground, arbuscular mycorrhizal fungi (AMF) are known to associate with *M. guttatus* [@bunn2009; @mcintosh2024].

Foliar symbionts can increase plant fitness under stressful conditions [@giauque2019; @aimone2023; @shankarnaik2019; @karasov2024]. Altitudinal gradients provide excellent natural laboratories to explore environmental effects on plant-microbe interactions and potential local adaptation of the microbiome. The *M. guttatus* species complex occupies a wide range of habitats in Western North America [@wu2008], including elevational gradients in the Sierra Nevada, CA where multiple closely related species occur sympatrically. This makes it an interesting system to examine the composition of FEF communities across species and elevation to determine if they align with broader ecological patterns. A study by Aguilar et al [-@salinasaguilar2022], revealed that in *M. laciniatus*, the bacterial community structure of roots and shoots and the substrate origin (soil, and moss) differed acrros a range of population and sites. Similarly, a study by Kazenel et al. [-@kazenel2019], found that the composition of root and leaf endophytes in cool-season, perennial grasses, varied differently with altitude and warming. Others have found that elevation and site are the strongest predictors of fungal taxa [@cordier2012]. However few studies have examined the effect of both host species and environmental variation on FEF community composition.

Here we use amplicon sequencing to explore how FEF communities vary among sympatric populations of *M. guttatus*, *M. laciniatus* and *M. nasutus* collected along replicated elevation gradients throughout the central Sierra Nevada, CA ([Fig. @fig-figure1]). We test the following questions: Q1) Are there differences in leaf functional traits among sympatric *Mimulus* species populations at similar elevations? Q2) Within each host species, how do leaf traits vary along the elevation gradient? Q3) Do host species and elevation jointly shape FEF richness and community composition, and do leaf functional traits predict FEF β-diversity? Q4) Within each host species, does FEF community composition change with geographic distance?  We include leaf thickness (LT), leaf punch strength (LPS), leaf mass per area (LMA), and anthocyanin content index (ACI) because these traits are known to influence the structure of FEF communities and play key roles in plant defenses against herbivores and pathogens [@saunders2010; @tellez2022; @gould2004; @aponterolon2024]. Additionally, we include leaf lobe index (LBI) to measure how distinct leaf shape is between the three *Mimulus* species and leaf lobing's potential effects on light exposure, temperature regulation, and hydraulic constraints within leaf tissue , which may, in turn, impact FEF symbionts [@ferris2015; @nicotra2011; @nobel2009]. We predicted that abundance, diversity,and richness would decline with increased elevation, and community composition of FEF would be more similar among the same sites (alpha diversity) than between sites (beta diversity) regardless of host species. Our results provide insights into how host genotype and environmental variation affect the ecological dynamics of plant-microbe interactions.

<!-- In addition to the central role of photosynthesis for explaining leaf shape, emphasizing specific leaf functional traits such as leaf thickness (LT), leaf mass per area (LMA), and leaf toughness can illustrate how mechanical properties impact plants' response to their abiotic and biotic environments [@oguchi2018]. Such characteristics are part of a leaf economic spectrum, with leaves at one extreme having short lifespans, high nitrogen content, low LMA, thin leaf blades, and thin cell walls. At the opposite end are long-lived leaves with low nitrogen content, high LMA, thick leaf blades and thick cell walls [@mason2015; @wright2004]. These traits not only determine the chemical, structural, and longevity attributes of leaves but influence their associated microbial communities [@saunders2010; @tellez2022].  -->

<!-- Leaves are essential structures for plants, playing a crucial role in various physiological processes such as photosynthesis, transpiration, and nutrient uptake [@oguchi2018; @fritz2018; @tsukaya2018]. They display various shapes driven by both genetic and environmental factors [@tsukaya2018; @tsukaya2005; @nicotra2011; @fritz2018]. Since photosynthetic leaves are crucial for plant growth and survival, it has been proposed that natural selection has shaped leaf functional traits in response to environmental conditions and ecological demands [@jump2005; @oguchi2018; @nicotra2010]. The leaf functional traits involved in the resulting leaf shape respond to three main stimuli: light, temperature modulation, and hydraulic constraints. Therefore, proposed explanations for the evolution of different leaf shapes have included: 1) thermoregulation to arid and hot environments, 2) hydraulic constraints and biomechanical constraints, 3) adaptations against plant enemies, and 4) adaptations to optimize light interception [@nicotra2011; @oguchi2018].  -->
<!-- These are not mutually exclusive, and the expected patterns are not always clear, for example, the optimal leaf width, the distance from mid-vein to the leaf margin, is driven evolutionarily by the environmental conditions experienced by species [@tsukaya2018].   -->

 <!-- For example, Wang et al. [-@wang2022] provide evidence of the favored genetic pathways for leaf shape differentiation in a comparative study  *Arabidopsis thaliana* which has narrow entire leaves, and close relative *Cardamine hirsute* which has deeply lobed leaves.  -->

# Materials and Methods

## Sample Collection

To examine the effects of geography and host species on FEF communities, we collected plant specimens during April-July 2021 and 2022 from populations of *M. guttatus*, *M. laciniatus*, and *M. nasutus* (syn. *Erythranthe guttata*, *Erythranthe laciniata*, and *Erythranthe nasuta*) across Stanislaus National Forest (SNF), Sierra National Forest (SINF) and Yosemite National Park (YNP), CA, USA ([@supptbl-1]).   We haphazardly selected sites close to the main roads based on the presence of a viable population with at least \~ 50 individuals per species. Samples collected from YNP were collected from non-wilderness areas on the side of the road. We determined population viability ensuring that they had individuals flowering or close to flowering stage. We collected between 5 - 20 individuals per species from a total of 25 sites (Table 1). We selected individuals that possessed healthy looking leaves and no visible signs of pathogen damage or senescence. At sites where two species were present, we collected individuals that were at least \~ 25 meters apart. We collected sample specimens by carefully uprooting the plant and placing into individual plastic bags (e.g., Ziploc®) and preserving in an ice chest until return to the field laboratory at the UC Merced Yosemite Field Station, YNP, CA, USA. Plant specimens were processed within 8 hrs of collection.

## Leaf trait measurements

To quantify phenotypic diversity among and within these species, we measured leaf thickness (LT), leaf punch strength (LPS), leaf mass per area (LMA), anthocyanin content index (ACI) and leaf lobe index (LBI) on all 309 field collected specimens, taking three measurements per trait from three haphazardly selected leaves per plant (one leaf for LBI). LBI was calculated in ImageJ [v1.52r; @schneider2012a] as the convex hull area minus the true leaf area divided by convex hull area, following Ferris et al. [-@ferris2015]. ACI was measured with a chlorophyll/anthocyanin content meter, LT with a digital micrometer gauge, and LPS (leaf toughness) with a digital force gauge via punch-and-die tests [@tellez2022]; LMA was determined from leaf disk punches dried at 60℃ for 48-72 hours before weighing. Full instrument models, vendors, and measurement parameters are provided in Supplementary Methods.

## Molecular Work

### Tissue preservation

We assessed FEF community composition using an amplicon sequencing approach. Upon completion of the leaf trait measurements, we surface-sterilized leaf strips (sequential ethanol and sodium hypochlorite washes) [@arnold2003; @tellez2022; @higgins2014] and preserved them in CTAB solution under sterile conditions at the UC Merced Yosemite Field Station. Full sterilization, cutting, and preservation protocol details are provided in Supplementary Methods.

### Amplicon sequencing

We extracted high-molecular-weight DNA using a CTAB:chloroform:isoamyl protocol combined with a solid-phase reversible immobilization (SPRI) bead step, adapted from Russo et al. [-@russo2022; @rohland2012; @liu2023] and optimized for fungal DNA from preserved leaf tissue [@aponterolon2023a]. We amplified the fungal ITS region using primers ITS1F [@gardes1993] and ITS2 [@white1990] with Illumina TruSeq adapters, following a two-step PCR amplification approach [@sarmiento2017; @uren2017; @lumibao2018], and included DNA extraction and PCR negative controls throughout to monitor contamination; none was detected visually or fluorometrically. Libraries were sequenced on the Illumina MiSeq platform with Reagent Kit v3 (2 × 300 bp) at the Duke Genome Sequencing and Analysis Core Facility (Durham, NC, USA). Full reagent, instrument, catalog, and reaction-volume details are provided in Supplementary Methods.

### Bioinformatic analyses

We assessed the quality of the reads using FastQc v0.12.1 [ v0.12.1; @andrews2010] and MultiQC [@ewels2016] tools. A total of 60,696,808 total ITS1 reads yielded from 343 (including 27 controls) libraries sequenced in two separate sequencing runs. The first sequencing event (177 samples and 15 controls) yielded 32,117,684 and the second (139 sample and 12 controls) 28,579,124 ITS1 reads. We tailored the open-source DADA2 bioinformatic pipeline for our data set [@callahan2016]. Based on our initial quality assessment, both forward and reverse reads were of low quality, with base calls deteriorating after 100 bp. We filtered our reads for ambiguous calls before removing the adapters by using `filterAndTrim` function and argument `maxN = 0` from the `dada2` package [v1.28.0; @callahan2016]. We removed forward and reverse primer adapters (and their reverse compliments) and eliminated reads shorter than 20 bp using the `cutadapt` tool [v4.6, @martin2011]. After removing ambiguous calls and forward and reverse primers, we assessed the quality of the reads again and saw slight improvement. We proceeded to apply stringent filter and truncation parameters to ensure quality of reads when assigning taxonomy. We filtered and truncated reads based on maximum expected errors (maxEE) rather than read length as it provides a reliable quality filtering [@edgar2015]. For this we set the arguments `trunQ = 2`, `maxEE = c(2,2)` for forward and reverse reads, and minimum read length of 50 bp with `minLen = 50` in the used the `filterAndTrim` function [@callahan2016]. These parameters eliminated all 151 libraries (139 samples and 12 controls) from our second sequencing event, leaving the 177 samples and 15 controls from the first sequencing event for downstream processing (see Results for final sample counts after subsequent quality-control steps). After this filter, we dereplicated reads with the `derepFastq` function and merged pairs using `mergePairs` functions with an overlap of 20 bp, minimum. We then inferred composition of the samples with `dada` function, which applies the DADA algorithm [@callahan2016; @rosen2012]. We removed chimeras via the "consensus" method with the `removeBimeraDenovo` function and ultimately, we used the `assignTaxonomy` function to assign taxonomy the amplicon sequence variants (ASV) referenced against the UNITE database [@abarenkov2023a]. After taxonomy assignment we used the `phyloseq` package [@mcmurdie2013] to create a phyloseq object for downstream analyses.

We used the `decontam` package [v1.20.0; @davis2018] to statistically determine which ASVs are likely contaminants based on their frequency in our samples and remove them using `prune_taxa` function from the `phyloseq` package [v1.44.0; @mcmurdie2013]. After which, we calculated the average read count found in DNA and PCR extraction controls, considered to be laboratory contaminants, and subtracted that from the samples' read counts. We then used custom scripts to remove any ASV that represented less than 0.1% of the abundance per sample on the assumption that it originates from contamination throughout handling of samples in the DNA and PCR processes. We removed singleton ASVs with the `prune_taxa` function [@mcmurdie2013]. We identified core taxa members at a 1% detection threshold and 50% prevalence in samples using the `core_members` function from the `microbiome` package [v.1.22.0; @lahti2012]. All post-quality bioinformatic steps were performed in *R* [v.4.4.1; @rcoreteam2024].

## Statistical Analyses

### Leaf traits

We checked for normality and homoscedasticity of the leaf traits measured using Shapiro-Wilk and Fligner-Killen tests from the `stats` package [@rcoreteam2024]. We established that the leaf functional trait data was not normally distributed and not homoscedastic. We then used non-parametric tests, the Wilcoxon Rank Sum test, to compare leaf functional trait means among species and sites to answer the first portion of Q1 and Q2. We used the `compare_means` and `stat_compare_means`functions from the `ggpubr` package [@kassambara2023] to perform these tests and properly visualize them. We adjusted *p* values to account for false discovery rates in multiple comparisons by using “BH” method [@benjamini1995]. We performed Principal Component Analysis (PCA) to understand patterns and relationships among leaf traits of host species. We used the `prcomp` function from the `vegan` package [@oksanen2022] compute the PCA analysis with variables ACI, LT, LPS, LMA, and LBI, all log-transformed. We used only complete raw leaf functional traits measurements to compute the PCA analysis (*n* = 504), samples with missing values were omitted.  All statistical analyses were performed in *R* programming language [v.4.4.1; @rcoreteam2024]. We present the log-transformed leaf functional trait data at the leaf level from a total of `r leaf_traits_noB |> filter(!Unique_ID == "FVWB_G002") |> distinct(Unique_ID) |> nrow()` plants: ACI (*n* = `r sum(!is.na(leaf_traits_noB$logACI))`), LT (*n* = `r sum(!is.na(leaf_traits_noB$logLT))`), LPS (*n* = `r sum(!is.na(leaf_traits_noB$logLPS))`), LMA (*n* = `r sum(!is.na(leaf_traits_noB$logLMA))`), LBI (*n* = `r sum(!is.na(leaf_traits_noB$logLBI))`). The FEF community data is presented from a subset of individuals at the plant/sample level (*n* = `r sum(!is.na(plant_traits$Observed))`).

### Community Diversity

To account for uneven sampling effort and over-representation of sequences, we normalized libraries by repeated rarefying, following Cameron et al. [-@cameron2021]. We determined a sequence depth of 750 by calculating Good's coverage and qualitative evaluation of libraries to determine a balanced coverage and breadth of samples [@schloss2024]. This approach allowed for a proportionate representation of observed sequences from host species and a robust characterization of random variation inherent in rarefaction of small libraries [@cameron2021; @schloss2024]. First, we randomly selected 136 samples out of the `r !is.na(ncol(ps_clean_3@otu_table@.Data))` viable samples that resulted from the bioinformatic pipeline (refer to the Results section below). The sample pool was reduced to 136 to match the theoretical reduction of one sample per site (*n* = 21). We generated 50 rarefied abundance matrices without replacement by using the `mirl` function from the `mirlyn` package [@cameron2021]. After which we were left with `r leaf_traits_noB |> filter(Unique_ID %in% colnames(asv_avgdist)) |> transmute(Unique_ID, Species) |> distinct() |> count() |> pull(n)` samples (*M. guttatus*: *n* = `r leaf_traits_noB |> filter(Unique_ID %in% colnames(asv_avgdist)) |> transmute(Unique_ID, Species) |> distinct() |> count(Species) |> filter(Species == "M. guttatus") |> pull(n)`, *M. laciniatus*: *n* = `r leaf_traits_noB |> filter(Unique_ID %in% colnames(asv_avgdist)) |> transmute(Unique_ID, Species) |> distinct() |> count(Species) |> filter(Species == "M. laciniatus") |> pull(n)`, *M. nasutus*: *n* = `r leaf_traits_noB |> filter(Unique_ID %in% colnames(asv_avgdist)) |> transmute(Unique_ID, Species) |> distinct() |> count(Species) |> filter(Species == "M. nasutus") |> pull(n)`). We then calculated α-diversity per sample as Hill orders, the observed richness (𝑞 = 0), the exponent of Shannon's entropy (𝑞 = 1), and the Inverse Simpson's Diversity (𝑞 = 2) by applying a modified version of the function `alphadivDF` [@cameron2021], which wraps the common diversity indices in `vegan` [@oksanen2022] (see custom script in Supplementary material). Finally, for beta diversity analyses, we performed a Hellinger transformation on the rarefied abundance matrices and calculated a Bray-Curtis dissimilarity matrix for each.

We calculated simple linear regressions and non-parametric tests, the Wilcoxon Rank Sum test, to understand how α-diversity changed in response to elevation (Q1) and (Q2). To facilitate our understanding of the effects of elevation on FEF α-diversity, we categorized elevation as "low" (\< 1219 m.a.s.l), "mid" (1220 - 1828 m.a.s.l.) and "high" (\> 1829 m.a.s.l.). To statistically compare the FEF community similarities within each host species across sites (Q3) we applied a distance-based Redundancy Analysis (dbRDA) on the Bray-Curtis dissimilarities which effectively describes underlying patterns of compositional differences [@anderson2017; @legendre1999; @mcardle2001]. We selected model terms by calculating Spearman's rho for all leaf traits (see below and Fig. S1) to understand which leaf functional traits were correlated with FEF richness, diversity and community composition. We informed our selection of leaf traits with results from the PCA (see below and Fig. 1) and selected those with the lowest correlation coefficient per host species: logLBI. Our initial dbRDA model consisted of terms logLBI, sampling date, site, and elevation (m). The leaf functional trait data, as well as elevation, was not randomized or subsampled to match rarefied dataset, the same values applied to all 50 rarefied matrices. For our initial model, we determined the variance inflation factor (VIF) of each term with function `vif.cca` and eliminated redundant ones: site and sampling date. We performed manual model selection by evaluating the marginal significance of constraining variables after 999 permutations with function `anova.cca` and argument `by` set to "terms" to assess significance for each term [@legendre2011; @legendre2012; @oksanen2022]. We corroborated homogeneous dispersion of variances with a permutational analysis of multivariate dispersion (PERMDISP) using the `betadisper` with parameter `type = "median"`, and `permutest` functions from `vegan`, the latter with 999 permutations [@oksanen2022]. We used a post-hoc Tukey's test to compare the differences in the dispersion of the FEF communities among sites and elevation zones. We used the `simper` function from `vegan` to discriminate which species contribute the most to compositional differences between groups [@oksanen2022].

We used generalized linear mixed models (GLMM) to determine the effect of host species, elevation, and leaf functional traits on the mean Bray-Curtis dissimilarity (β-diversity) of FEF communities (Q3). Leaf functional traits with missing values were imputed with the median value for the trait. We used the `lme` function from the `nlme` package [v3.1-165; @pinheiro2024; @pinheiro2000] to fit the GLMM models. We established leaf functional traits, elevation and host species as fixed terms and site as a random effect. We modeled variance structure for site with the `varIdent` argument. We manually compared and selected models based on Akaike Information Criterion (AIC) with a penalty of 2 degrees of freedom (ΔAIC) with the `model.sel` function from the `MuMIn` package [@barton2023]. We selected the best-fit model based on the lowest value obtained. All models are based on restricted maximum likelihood estimation (REML).
 
To assess correlations between pairwise FEF community dissimilarity and the geographical distance matrix per host species (Q4), we computed Mantel tests with Spearman's rho and 999 permutations using the `mantel` function [@oksanen2022]. For this test, we opted for a less taxing computational approach and rarefied sequences with the same parameters as before and calculated Hellinger transformations with the `avgdist` function [@oksanen2022]. We then calculated the Bray-Curtis dissimilarity with `vegdist` [@oksanen2022]. For the geographical distances between sites, we used `distm` function with the Vicenty (ellipsoid) method from the `geosphere` package [@hijmans2022].



# Results

## Q1) Host species differ in leaf functional traits at shared elevations

We found evidence of both interspecific and intraspecific elevational shifts in leaf functional traits among sympatric *Mimulus* species, addressed in Q1 and Q2 below, respectively ([Fig. @fig-figure2] and Figs. S5-S9). Our PCA analysis showed how leaf functional traits varied among sympatric *Mimulus* species populations at similar elevations (Q1). We plotted leaf functional traits according to species groups on the PCA axes to show how the variance in the complete data set is explained by PC1 (42.57%) and PC2 (23.27%) ([Fig. @fig-figure2]). The PCA showed correlations between ACI, LT, LPS, and LMA as loadings that tracked along PC1 towards positive values ([Fig. @fig-figure2]). We observed that LBI loading was orthogonal along PC2 to the other traits, indicative of low correlation. We noted distinct groupings by species along PC2 such that *M. laciniatus*, the most lobed species, was distinct from *M.guttatus* and *M. nasutus*. The latter two overlapped considerably along PC1 and PC2 ([Fig. @fig-figure2]), but species means still differed significantly along PC1 (ANOVA, *F*~2,~ ~501~ = 15.55, *p* < .001): *M. guttatus* had a positive mean PC1 score while *M. laciniatus* and *M. nasutus* had similar, negative mean scores, indicating that the ACI/LT/LPS/LMA loadings underlying PC1 also weakly separate *M. guttatus* from its congeners, in addition to the LBI-driven separation of *M. laciniatus* along PC2.

<!-- Q1) Are there differences in leaf functional traits among sympatric *Mimulus* species populations at similar elevations?  -->

At shared elevations, leaf functional traits frequently showed interspecific differences, though the magnitude and direction of differences varied across elevation zones (Figs. S5B-S9B). At mid-elevations, where all three species co-occurred, *M. laciniatus* exhibited distinct trait values compared to *M. guttatus* and *M. nasutus*, including higher leaf mass per area (LMA, *p* \< .001; Fig. S5B), lower anthocyanin content (ACI, *p* \< .0001; Fig. S6B), and greater lobing (LBI, *p* \< .0001; Fig. S7B). At high elevations, *M. nasutus* had higher LMA and higher ACI than both congeners (LMA: *p* \< .0001; Fig. S5B; ACI: *p* \< .0001; Fig. S6B), consistent with its steeper elevational increase in LMA reported above (Q2), while *M. guttatus* had thinner leaves than *M. nasutus* (LT, *p* \< .0001; Fig. S9B). Notably, LBI showed minimal overlap among species across all elevations (Fig. S7B), reflecting strong species-specific morphological constraints.

## Q2) Trait-elevation relationships are species-specific

<!-- Q2) Within each host species, how do leaf traits vary along the elevation gradient?  -->

Trait-elevation relationships varied by species and trait type (Q2). Across the entire community, LMA, LPS, and LT increased weakly but significantly with elevation (LMA: *R²~adj~* = .041, *p* < .001; LPS: *R²~adj~* = .004, *p* = .032; LT:*R²~adj~* = .013, *p* < .0001; Figs. S5A, S8A, S9A). However, this pooled trend masked substantial species-specific differences. Per-species regressions of each trait on elevation showed that *M. nasutus* increased significantly in all three traits (LMA: *R²~adj~* = .278, *p* < .0001; LPS: *R²~adj~* = .199, *p* < .0001; LT: *R²~adj~* = .067, *p* = .002), *M. laciniatus* increased significantly in LMA and LT but not LPS (LMA: *R²~adj~* = .044, *p* = .01; LT: *R²~adj~* = .047, *p* < .0001; LPS: *p* = .12), and *M. guttatus* showed no significant elevational increase in LMA or LPS (*p* = .77 and *p* = .25, respectively) and a significant *decrease* in LT with elevation (*R²~adj~* = .01, *p* = .023). Categorical analyses were consistent with these trends: *M. laciniatus* maintained consistently high LBI across elevations (Fig. S7B), while *M. guttatus* showed increased LPS at mid-elevations (*p* < .0001; Fig. S8B) and greater LT differentiation at higher elevations (*p* \< .0001; Fig. S9B). In contrast, ACI showed no community-wide elevational trend (*R²~adj~* \< -.000, *p* \< .0001; Fig. S6A), but *M. laciniatus* exhibited lower ACI at higher elevations compared to congeners (*p* \< .01; Fig. S6B).

## Q3) Host species and elevation jointly shape FEF richness and community composition; leaf functional traits predict β-diversity

<!-- Q3) Do host species and elevation jointly shape FEF richness and community composition, and do leaf functional traits predict FEF β-diversity?  -->

We obtained 5,082,229 reads representing 726 ASVs from 174 samples after processing samples through the DADA2 pipeline. The raw reads obtained were composed of 26.81% Ascomycota, 71.53% Basidiomycota, \<0.05% Chytridiomycota, \<0.5% Mortierellomycota, \<0.03% Olpidiomycota, 0.01% Rozellomycota, \<0.001% Aphelidiomycota, and 1.19% unidentified. After decontamination, and removal of singletons we retained 4,856,220 reads representing 231 ASVs from 157 samples composed of 26% Ascomycota, 73% Basidiomycota, 0.01% Chytridiomycota, \<0.1% Mortierellomycota, 0.03% Olpidiomycota, \<0.002% Rozellomycota and \<1.0% unknown reads ([Fig. @fig-figure3] and Fig. S2).

Host species did structure FEF community composition even among populations at similar elevations: composition differed significantly by host species and by elevation zone (see below), and communities were most distinguishable by host species at high elevations specifically, where community overlap between species was lowest.

**FEF α-diversity declined with elevation, but responses were species specific.** We observed significant interspecific differences in α-diversity across all Hill numbers (𝑞 = 0, 1, 2), with the sole exception of the *M. laciniatus* - *M. nasutus* contrast at 𝑞 = 0 (Fig. S4A-S4C) (Q3). FEF α-diversity decreased significantly with elevation across all Hill numbers:  (𝑞 = 0) (*R^2^~adj~* \< .01, *F*~1,~ ~4198~= 16.69, *p* \< .001), the exponent of Shannon's entropy (𝑞 = 1) (*R^2^~adj~* \< .01, *F*~1,~ ~4198~= 37.91, *p* \< .001), and the Inverse Simpson's Diversity (𝑞 = 2) (*R^2^~adj~* \< .01, *F*~1,~ ~4198~= 23.73, *p* \< .001). Species-specific responses differed: *M. nasutus* showed declining diversity with elevation ([Fig. @fig-figure4]), while *M. laciniatus* exhibited increased diversity. *M. guttatus* displayed mixed trends, with elevated Hill *q* = 0 diversity but decreased diversity at higher Hill orders. We observed a similar pattern in beta diversity among elevation sites (Fig. S4D -S4F). We saw no differences in beta diversity between low and mid elevation sites for Hill order 2 (𝑞 = 2, *p* \> .05, Fig. S4F).

**Host species and elevation jointly structured FEF community composition.** FEF communities differed significantly among host species and elevation zones. The best-fit model revealed that 22% of the overall variance in FEF communities was accounted for by log-transformed LBI, and the interaction of host species and elevation, all of which were significantly correlated  (*p* \< .001) with FEF communities in their host species ([Fig. @fig-figure5] and [@supptbl-3]). We observed that the first axis (dbRDA1) explained 51.7% and the second axis (dbRDA2) explained 15.9% of the constrained variance ([Fig. @fig-figure5]) which support that host species and elevation interactions structure FEF communities.  Communities showed greater overlap at low elevations but became distinct at high elevations  ([Fig. @fig-figure5]).

FEF community composition differed significantly among host species (*F*~2,~ ~4197~= 320.3, *p* < .001) and among elevation zones (*F*~2,~ ~4197~= 228.1, *p* \< .001), with all pairwise comparisons significant by post-hoc Tukey’s test (⍺ = 0.05); the interaction between log LBI and elevation zone showed the same pattern (*F*~2,~ ~4192~= 268.1, *p* < .001). PERMDISP tests showed that within-group dispersion also differed for most of these same comparisons, so dispersion is a plausible contributor to some of the apparent compositional differences. However, dispersion was not universally different — it was homogeneous for *M. guttatus* across all elevations, for *M. laciniatus* (LOW/MID) vs. *M. guttatus* (HIGH), and for *M. nasutus* (MID/HIGH) vs. *M. laciniatus* (MID) (Tukey’s test, ⍺ = 0.05) — and the same host-species and elevation effects on composition were independently recovered by the dbRDA above and by the GLMM below, neither of which relies on the PERMDISP test. This convergence across independent analyses indicates that dispersion differences are unlikely to fully explain the observed community structure.

**Leaf functional traits, not elevation, predicted β‑diversity.** GLMMs identified leaf functional traits, not elevation, as primary predictors of mean β-diversity (Model 4, Table 1). Anthocyanin content (ACI) showed the strongest negative association (estimate = -0.078, *t*(4200) = -32.48, *p* \< .001), followed by leaf thickness (LT; estimate = -0.051, *t*(4200) = -9.51, *p* \< .001) and leaf mass per area (LMA; estimate = 0.013, *t*(4200) = 3.74, *p* \< .001). Simple linear regressions showed that *M. nasutus* (*R^2^~adj~* = .46, *F*~1,~ ~748~= 642.3, *p* \< .001) and *M. guttatus* (*R^2^~adj~* \< .01, *F*~1,~ ~1848~= 38.3, *p* \< .001) exhibited elevational declines in β-diversity, while *M. laciniatus* showed no elevational trend  (*R^2^~adj~* \< .01, *F*~1,~ ~1598~= 1.84, *p* = .176, [Fig. @fig-figure6]).

## Q4) Geographic distance predicts FEF turnover in M. laciniatus and M. nasutus but not M. guttatus

<!-- Q4) Within each host species, does FEF community composition change with geographic distance? -->


 Finally (Q4), we detected clear distance-decay in FEF mean β‑diversity turnover in *M. laciniatus* (*r* = 0.27, *p* = .01) and *M. nasutus* (*r* = 0.29, *p* < .01, [Fig. @fig-figure7]), consistent with isolation by distance and/or spatially structured environmental filters acting on foliar fungi. In contrast, *M. guttatus* (*r* = -0.004, *p* = .48) showed no evidence of distance-decay ([Fig. @fig-figure7]), consistent with broader dispersal of its dominant endophytes, greater host gene flow that weakens population structure and geographic filtering, or stronger local environmental/host-trait filtering that overwhelms spatial effects. Distance-decay in two selfing species but not in *M. guttatus* suggests that dispersal limitation and host spatial structure shape FEF communities, with species-specific balances between geographic and environmental filtering.

 
<!--  Effect sizes are modest, as expected for microbial systems, but they indicate that dispersal limitation and geography measurably shape endophyte assembly in these hosts. -->

<!-- In contrast, M. guttatus  (r ≈ 0), suggesting either broader dispersal of its dominant endophytes, weaker spatial population structure of the host that reduces geographic filtering, or stronger local environmental/host-trait filtering that overwhelms distance effects. -->

<!-- The species contrast aligns with host life history: both M. laciniatus and M. nasutus are predominantly selfing and often exhibit stronger population genetic structure and habitat specialization than M. guttatus. Tighter host genetic/spatial structuring could reinforce geographic sorting of symbionts, whereas the more outcrossing M. guttatus may experience higher gene flow and/or associate with more widespread fungal taxa. -->

<!-- Integrating with other results, the spatial signal in M. nasutus complements its elevational decline in β-diversity (Fig. 6), implying both space and environment contribute to turnover. For M. laciniatus, geography mattered while elevation did not, pointing to spatial processes operating independently of the elevational gradient. Conversely, M. guttatus showed elevational but not spatial turnover, consistent with environmental filtering rather than dispersal limitation. -->

<!-- Caveats and next steps: Mantel tests conflate space with environment; partial Mantel or multi-predictor models (e.g., MRM) that include elevation and climate, or spatial eigenvector approaches (dbMEM/PCNM) would help partition spatial vs environmental effects. Uneven sample sizes and inter-site distances can also influence Mantel power; targeted, within-region sampling could test for distance-decay at finer scales. -->

## Additional finding: a consistent, prevalent core of foliar taxa across hosts and elevations

After rarefaction of sequences, we were left with 84 samples in which we found the most prevalent taxa: *Vishniacozyma victoriae* (Basidiomycota, ASV_1), *Cladosporium herbanum* (Ascomycota, ASV_2) and *Cladosporium* spp. (Ascomycota, ASV_7), *Dyszogia patagonica* (ASV_3), *Filobasidium chernovii* (Basidiomycota, ASV_5), and *Alternaria tenuissima* (Ascomycota, ASV_8) (Fig. S2). These taxa were consistently detected across host species and elevation zones (see Discussion).




# Discussion
In this study, we examined how leaf functional traits and foliar endophytic fungal (FEF) communities vary among three sympatric *Mimulus* species across elevation gradients, and whether elevation, host identity, and leaf traits jointly structure FEF richness, diversity, composition, and spatial turnover. Below, we discuss our findings in the same order as our four guiding questions: interspecific differences in leaf functional traits at shared elevations (Q1), species-specific trait-elevation relationships (Q2), how host species and elevation jointly shape FEF richness and community composition (Q3), and geographic distance-decay in FEF turnover (Q4), followed by the consistent dominant taxa observed across sites.

<!-- In an experimental study, McIntosh et al. [-@mcintosh2024], observed that coastal perennial and montane annual ecotypes of *M. guttatus* harbored distinct fungal communities, with the coastal perennial ecotype being more colonized by AMF than the montane annual ecotype. The authors suggested that population divergence in life history and related traits shapes plant responsiveness to soil microbes and the specificity of their associations with root endophytic fungi, such as AMF [@mcintosh2024].  -->

## Q1) Host species differ in leaf functional traits at shared elevations

 <!-- Auxin-related pathways have been well documented as one of the main mechanisms for the expansion of leaf lamina [reviewed in @xiong2019].   -->
We found clear interspecific differentiation in leaf functional traits among sympatric *Mimulus* species, with leaf lobing (LBI) providing a particularly strong axis of separation among hosts. The Principal Component Analysis (PCA) ([Fig. @fig-figure2]) illustrated correlations between leaf functional traits while demonstrating the distinct groupings of *M. laciniatus*, *M. guttatus*, and *M. nasutus* based on LBI differences compared to the rest of the traits: ACI, LMA, LPS and LT.

## Q2) Trait-elevation relationships are species-specific

Across the pooled dataset, LMA, LPS and LT were weakly but significantly positively correlated with elevation at the genus level; however, per-species regressions showed that this trend was not shared uniformly among species (see Results). *Mimulus nasutus* showed the clearest and most consistent elevational increase across all three traits, *M. laciniatus* increased significantly in LMA and LT but not LPS, and *M. guttatus* showed no significant elevational increase in LMA or LPS and, notably, a significant elevational *decrease* in LT. *Mimulus laciniatus* is the most lobed species ([Fig. @fig-figure2] and Fig. S7) [@ferris2015; @ferris2016; @tataru2023; @sexton2013a], and thrives at higher elevation. For serrated/lobed leaves, it has been observed that physiological adaptations in the leaf margin are a response to the average air temperature, but this is not a simple relationship [@tsukaya2018; @nicotra2011].

The elevational increase in leaf lobing suggests that *M. laciniatus* leaf morphology tracks microhabitat rather than ambient elevational temperature. *M. laciniatus* is distinguished from its congeners by its specialization on exposed granite rock-outcrop habitats with shallow, well-drained soils and high solar exposure [@ferris2014; @ferris2014a], which can create locally hot, dry growing conditions largely independent of — and sometimes opposite to — the average cooling and increased snowmelt that typically accompany elevation gain in this system; *M. guttatus* and *M. nasutus* more typically occupy moister streamside and seep habitats within the same elevation bands [@wu2008]. Reciprocal transplant work between *M. guttatus* and *M. laciniatus* similarly found the strongest divergent, habitat-mediated selection on quantitative traits during a drought year rather than a wet year, consistent with outcrop microhabitat, rather than elevation *per se*, driving trait divergence between these species [@tataru2023]. The observed increase in LBI with elevation, previously documented in *M. laciniatus* by Love and Ferris [-@love2024], is therefore more plausibly linked to this outcrop microhabitat than to ambient elevational cooling: lobed leaves are expected to have increased hydraulic efficiency and an advantage in locally hot, dry microsites by reducing the leaf boundary layer, an area of still air adjacent the leaf's abaxial surface [@nobel2009; @oguchi2018], which affects a leaf's gas and heat exchange [@schuepp1993; @nobel2009]. We restrict this interpretation to *M. laciniatus* because it is the only species with a documented elevational increase in LBI and the only one specializing on outcrop microhabitats; *M. nasutus* is not a notably lobed species in our data ([Fig. @fig-figure2]), and neither @wu2008 nor @ferris2014 characterize *M. nasutus* habitat as comparable to *M. laciniatus*'s outcrop specialization.

The genus-wide increase in LMA, LT and LPS with elevation is harder to attribute to a single mechanism from our data alone. A straightforward "reduce water loss" explanation is difficult to reconcile with this system, where higher elevations receive more snowmelt and cooler ambient temperatures rather than drier or hotter conditions, if anything, water availability likely increases with elevation here. Structural trait increases at elevation could instead reflect other, non-mutually-exclusive pressures such as greater UV exposure, wind, or diurnal freeze-thaw cycling, but we did not measure these variables directly and cannot distinguish among them with the present data. This pattern was driven primarily by *M. nasutus* and, to a lesser extent, *M. laciniatus*; *M. guttatus* showed no comparable increase in LMA or LPS and an elevational decline in LT, suggesting either weaker selection for these traits at elevation in this species or a distinct, uncharacterized strategy for coping with elevational conditions.

Disentangling environmental and genetic drivers of leaf trait variation will require various experimental approaches such as common gardens (as in Ferris and Willis, -@ferris2018; Dong et al -@dong2024; Love and Ferris -@love2024; Love and Ferris @love2025; Love et al @love2026; ) to fully answer questions about the functional significance of leaf traits for local adaptation to different elevations. Experimental manipulation with *Mimulus* is relatively easy due to its short generation time, high fecundity, self-compatibility, and ease of greenhouse propagation [@society2019; @wu2008]. The same tractability makes it possible to extend these designs beyond leaf traits: manipulating host genotype and functional traits directly, rather than inferring their effects from natural variation, would let future work test whether host genetic loci shape FEF community assembly and how these communities in turn affect host performance under elevational stress. Because these traits can act as host filters on microbial colonization, we next consider whether FEF diversity shows parallel elevational patterns within species.

## Q3) Host species and elevation jointly shape FEF richness and community composition; leaf functional traits predict β-diversity

**Within-species symbiont α-diversity declined with elevation.** Foliar endophytic fungi α-diversity declined with elevation ([Fig. @fig-figure4]), consistent with broader macroecological patterns of biodiversity [@sabatini2018; @kraft2011; @villacampa2019; @jimenez-hernandez2020], but the strength and direction of this pattern differed among host species. Alpha diversity declined with elevation for *M. nasutus*, *M. laciniatus* exhibited the opposite, while *M. guttatus* displayed mixed trends ([Fig. @fig-figure4]). The observed decline in α-diversity appears to be driven by *M. nasutus'* decrease and *M. guttatus*' unchanged diversity levels with increased elevation. This species specificity motivates examining not only diversity patterns but also how elevation, host identity, and leaf traits structure the composition of FEF communities.

**Host species and elevation jointly structured FEF community composition.** Consistent with the idea that both abiotic gradients and host identity filter symbiont communities, we found that host species, elevation, and LBI were associated with differences in FEF community composition. Our dbRDA results ([Fig. @fig-figure4]) indicate that host species, elevation, and LBI account for 17% of the variance in FEF community composition. Subsequent PERMDISP analyses also support significant differences in FEF community composition between host species at low, mid, and high elevations. However, when we modeled mean β-diversity directly, leaf functional traits rather than elevation emerged as the strongest predictors, prompting a closer look at how analytical approaches shape inference.

Although ordination-based results highlighted elevation and host identity, GLMM results showed that ACI, LT, and LMA are more direct predictors of mean β-diversity than elevation *per se* ([Fig. @fig-figure6]). Apparent discrepancies likely arise from model specification: the GLMM explicitly modeled site-level variance (Table [@tbl-table1]), whereas the dbRDA did not partition site from elevation, potentially introducing unaccounted site effects.

## Q4) Geographic distance predicts FEF turnover in M. laciniatus and M. nasutus but not M. guttatus

Within species, spatial structure also contributed to FEF turnover: community dissimilarity increased with geographic distance in *M. laciniatus* and *M. nasutus* according to our Mantel tests, providing evidence in support of spatially driven differences in FEF community composition. It is interesting to note that the most dissimilar FEF communities due to geographical distance are in the two species that have self-fertilizing mating systems, *M. laciniatus* [@ferris2014] and *M. nasutus* [@brandvain2014]. This could indicate that mating systems and host species' genetic structure could play a role in structuring FEF communities. Self-fertilizing species tend to have reduced within population genetic diversity and increased between population divergence due to high levels of inbreeding and reduced gene flow [@sweigart2003;@hartfield2017; @sexton2016].  Our findings overall support the idea that distinct FEF communities are structured by the interplay of host specific leaf functional traits and geographic distance. To place these patterns in a broader microbial-ecology context, we next compare our results with elevation-gradient studies in other plant systems and tissue types.

Comparisons with other elevation-gradient studies suggest that the relative importance of elevation, host identity, and tissue type can vary substantially across systems. Patterns of diversity and community composition in microbial ecology are often constrained by both biotic and abiotic factors. For example, Bowman and Arnold [-@bowman2021] showed distance decay in ectomychorrizal fungi (EMF) and foliar endophytes, but drivers differed for each: EMF were constrained by dispersal limitation, whereas foliar endophytes were constrained environmental conditions. Meanwhile, in an experimental setup, Kivlin et al., [-@kivlin2022] reported that host species (alpine grasses) was a stronger predictor than elevation of α-diversity and community composition of leaf endophytes. In contrast, root endophyte communities responded to both host species and elevation [@kivlin2022]. It is possible that our results differ due to the different phenologies and tissue types of herbaceous and gramineous plants. Similarly, Kezenel et al. [-@kazenel2019] found greater change in leaf endophytes due to altitude and warming when compared to root colonizing fungi, but the direction and magnitude of responses varied among host species and fungal functional groups. A major difference in this study is the low ASV count and the use of multiple rarefied data sets compared to Kazenel et al. [-@kazenel2019]. A study by Cordier et al. [-@cordier2012] focused on the fungal phyllosphere in European beech along an elevation gradient, and found that climatic variables, especially temperature, showed the strongest correlations with fungal community dissimilarities. While the phyllosphere of beech varies widely, Cordier et al. [-@cordier2012] also found a strong affinity of fungal taxa to elevation and site, supporting regional spatial structure. An important distinction is that they focused on the outer and inner phyllosphere, hence observing patterns that reflected the outer leaf dynamics. These may have been more susceptible to climatic factors, as opposed to inner leaf dynamics that we explored in our study. Other key differences are the host species phenology and functional leaf traits of European beech compared to herbaceous  *Mimulus* spp.

## *Additional finding: consistent taxa underpin FEF communities across elevations*

Despite turnover with elevation and distance, several taxa were consistently prevalent across sites and hosts, suggesting a core set of widespread foliar symbionts associated with in Sierran *Mimulus*. The widespread detection of *V. victoriae* across the elevation gradient is particularly notable given its documented associations with cold environments and its proposed functional roles. Additionally, we detected the consistent prevalence of *Cladosporium herbanum*, and *Cladosporium* in across all sites and samples.
 <!-- *Cladosporium* spp. is a well-documented saprophytic genus that occurs in senescing or dead leaves of herbaceous and woody plants [@samson2004; @schubert2007].  -->
 The basidiomycetous yeast *V. victoriae* (formerly *Cryptococcus victoriae*) is also a well-known environmentally abundant fungus capable of causing respiratory issues in humans [@rush2023]. It was first isolated in the Antarctic [@montes1999] and has since been detected worldwide [@demenezes2019]. It is proposed that through the production of various bio-active compounds, *V.victoriae* can contribute to plant growth and ecological adaptation to cold environments [@vujanovic2021; @buzzini2018]. 
 <!-- According to Vujanovic [-@vujanovic2021] and  Ogaki et al. [-@ogaki2020], no antagonism has been detected between *V. victoriae* with other yeasts, or plant pathogens.  -->
 Despite potential respiratory health detriments, *V. victoriae* has been utilized in agricultural settings for the post-harvest control of fruit diseases [@lutz2020]. It thrives at low temperatures (15 °C), but it is known to tolerate a variety of environmental conditions, and lacks a  polysaccharide capsule, which is thought to contribute to its lack of pathogenicity [@rush2023]. Its applications in wheat agriculture suggest that kernel weight is influenced by *V. victoriae*'s coexistence with other plant acquired endophytic fungi [@vujanovic2021]. Its presence might serve as an indicator of wheat's kernel resistance to pathobiota [@lutz2020]. We need further quantitative studies to confirm the existence of cold-adapted microbial taxa and their associated hosts [@marian2022]. The presence of *V. victoriae* in our samples might be indicative of its potential role in the local adaptation of *Mimulus* to cold and high elevation environments. Future surveys using fresh or rapidly frozen tissue rather than room-temperature CTAB preservation may improve DNA extraction yields and taxon recovery, sharpening inference on which taxa are truly core versus under-detected in our sampling.

# Conclusions

Overall, our results show variation in leaf functional traits and FEF diversity, composition, and spatial turnover across elevation in three sympatric Mimulus species. From our results, PCA and trait-elevation trends highlight host filtering; dbRDA shows host-elevation interactions; GLMMs point to leaf functional traits as strong predictor over elevation for mean β-diversity; Mantel tests reveal distance-decay where dispersal and host structure limit mixing.

Across mountain landscapes, foliar symbiont communities emerge from the joint action of three broad ecological filters: host functional traits, abiotic gradients, and spatial processes. In our system, trait differences among closely related hosts (Q1-Q2), a community-wide decline in α-diversity with elevation (Q3), trait-linked shifts in β-diversity (Q3), and distance-decay in selfing species (Q4) collectively point to an overarching assembly framework in which: (1) Host trait filtering (e.g., leaf thickness, toughness, lobing) sets the local rules of colonization and persistence within leaves; (2)Environmental gradients (e.g., temperature, water availability indexed by elevation) modulate trait distributions and community overlap across sites; (3) Spatial processes (e.g., dispersal limitation, isolation by distance) and host population structure determine the scale and strength of symbiont turnover.

These findings motivate several general ecological questions that extend beyond the focal hosts: (a) What is the relative importance of host trait filtering versus abiotic filtering and how does it shift across spatial scales, plant tissues, and seasons?; (b) How and when does host genetic structure amplify geographic filtering of symbionts? (c) What functional roles do cosmopolitan, cold-associated taxa play in plant performance across climate gradients, and how stable are these roles over time?; (d) How can trait-based and spatially explicit models be integrated to forecast microbiome assembly and function under climate change? (e) To what extent are these assembly rules conserved across plant lineages and fungal guilds, and where do they break down?

Addressing these questions will require synthesis across comparative field surveys, manipulative experiments, and trait-informed community models (see Discussion for the specific common-garden, genotype-manipulation, and tissue-preservation directions this system enables). By situating foliar symbionts within a trait-environment-space framework, future work can move from pattern to mechanistic models that quantify how plant traits and geography govern symbiont assembly, and how those communities feed back to influence plant resilience as climates change and distributions shift.


# Author Contributions
B.A.R, K.G.F. and S.A.V. designed research study. B.A.R. conducted field surveys, laboratory work, and collected data. B.A.R. processed the data and performed statistical analyses. B.A.R. wrote the manuscript with input from all authors. All authors gave final approval for publication. 


# Acknowledgements

This work was supported by the National Institute of General Medicinal Sciences of the National Institute of Health (NIH) under award number R35GM138224 to KGF. We thank the Yosemite National Park Service for permit support (YOSE-2021-SCI-0033, YOSE-2022-SCI-0051), and Breeanne Jackson at Yosemite Field Station (doi:10.21973/n3v36c) for logistical support and providing accommodation. We thank the USDA Forest Service for providing access to the Sierra National Forest and Stanislaus National Forest (Botanical Collection Permit: 031528). We thank Caroline, M. Dong, Fidel J. Machado Perez, Elizabeth MacDougal, Rachael Dennis, and Lissette Montano for field assistance. This research was supported in part using high performance computing (HPC) resources and services provided by Information Technology at Tulane University, New Orleans, LA. Lastly, we would like to acknowledge the original inhabitants of the unceded land on which our research was conducted, the Southern Sierra Miwuk Nation, Bishop Paiute Tribe, Bridgeport Indian Colony, Mono Lake Kutzadikaa, North Fork Rancheria of Mono Indians of California, Picayune Rancheria of the Chukchansi Indians and the Tuolumne Band of Me-Wuk Indians. Their culture and stewardship remain an integral part of the land. As scientists we strive to take responsibility for the impacts of colonialism in our field and move forward with respect and support of indigenous movements and knowledge.

# Conflict of Interest Statement

The authors declare no competing interests.

# Data Accessibility Statement
The genomic data that support the findings of this study will be submitted to NCBI-GenBank upon acceptance of this manuscript. The manuscript and code for reproducibility is available from corresponding author's GitHub:[https://github.com/jibarozzo/endophyte_mimulus.git](https://github.com/jibarozzo/endophyte_-leaf-traits_mimulus.git)
#TODO Submit to Zenodo and SRA Bioproject

# Benefit Sharing Statement
#TODO
# Author Contributions
B.A.R, K.G.F. and S.A.V. designed research study. B.A.R. conducted field surveys, laboratory work, and collected data. B.A.R. processed the data, performed statistical analyses, and maintains code for reproducibility. B.A.R. wrote the manuscript with input from all authors. All authors gave final approval for publication.

# References

::: {#refs}
:::

# Figures

```{r, themes}
#| echo: false
#| eval: true
#| tidy: true
#| warning: false

#ggplot themes
# Theme for leaf trait and diversity plots: discrete
theme_lfspp_discrete <- theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(linetype = "blank", fill = NA),
    legend.text = element_text(face = "italic"), #Make legend text italic
    axis.text.x = element_blank()
  )

# Theme for leaf trait plots: continuous
theme_lfspp_continuous <- theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    panel.border = element_rect(linetype = "blank", fill = NA),
    legend.text = element_text(face = "italic"), #Make legend text italic
    axis.text.x = element_text(size = 10),
    legend.position = "bottom"
  )

# Theme for species by diversity  plots: discrete
theme_sppdiv_discrete <- theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "italic", size = 10),
    panel.border = element_rect(linetype = "blank", fill = NA),
    legend.text = element_text(face = "italic", size = 12), #Make legend text italic
    axis.text.x.bottom = element_blank(),
    axis.title.x = element_blank()
  )

# Theme for elevation plots: discrete
theme_elevdiv_discrete <- theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "italic", size = 10),
    panel.border = element_rect(linetype = "blank", fill = NA),
    axis.text.x.bottom = element_blank(),
    axis.title.x = element_blank()
  )

# Theme for leaf trait and diversity  plots: continuous
theme_lfdiv_continuous <- theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold.italic", size = 10),
    panel.border = element_rect(linetype = "blank", fill = NA),
    legend.text = element_text(face = "italic"), #Make legend text italic
    axis.text.x = element_text(size = 10)
  )
```


## Figure 1

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 6
#| label: fig-figure1
#| fig-cap: "Map of *Mimulus* spp. sampling sites in the Sierra Nevada, California, USA. A) Map of the Sierra Nevada region showing the location of the three species sampled. The color gradient represents the elevation gradient from low (yellow) to high (purple)."

# Map of sampling sites

#register_google(key ='SECRET_KEY', write = TRUE)

map_latlong <- read_excel(
  "data/field_data/Mimulus_CH2_Field_Survey.xlsx",
  sheet = "clean_site_lat_long"
)

map_latlong <- plant_traits |>
  select(Site, Species, Longitude, Latitude, Elevation_m) |>
  filter(
    !Site == "FVWB" &
      !Site == "LYFB" &
      !Site == "TRTB" &
      !Species == "M. bicolor"
  )

mappops <- get_map(
  location = c(
    lon = mean(map_latlong$Longitude),
    lat = mean(map_latlong$Latitude)
  ),
  zoom = 9,
  maptype = "terrain",
  scale = 4
)
map <- ggmap(mappops, extent = "device", darken = 0) +
  geom_point(
    data = map_latlong,
    aes(
      x = Longitude,
      y = Latitude,
      size = Elevation_m,
      fill = Species
    ),
    shape = 21,
    alpha = 0.1
  ) +
  scale_fill_manual(
    labels = c("*M. laciniatus*", "*M. nasutus*", "*M. guttatus*"),
    values = met.brewer(
      "Isfahan2",
      n = 4,
      type = "discrete",
      direction = c(-1)
    )
  ) +
  scale_color_manual(
    labels = c("*M. laciniatus*", "*M. nasutus*", "*M. guttatus*"),
    values = met.brewer(
      "Isfahan2",
      n = 4,
      type = "discrete",
      direction = c(-1)
    )
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    legend.text = element_markdown(size = 10),
    panel.border = element_rect(linetype = "blank", fill = NA),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  ) +
  ylab("Latitude") +
  xlab("Longitude") +
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    height = unit(0.7, "cm"),
    width = unit(0.7, "cm"),
  ) +
  guides(
    fill = guide_legend(title = "Species", override.aes = list(alpha = 0.7)),
    size = guide_legend(
      title = "Elevation (m a.s.l.)",
      override.aes = list(alpha = 1)
    )
  )

map
ggsave(
  "figures/figure1.jpg",
  plot = map,
  units = "in",
  width = 6,
  height = 6,
  dpi = 600
)
```

## Figure 2

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 6
#| label: fig-figure2
#| fig-cap: "The Principal Component Analysis (PCA) shows how leaf functional traits are correlated. The PCA analyses shows correlations between ACI, LT, LPS, and LMA as loadings track along PC1 towards positive values, while leaf lobing correlates positively with M. laciniatus. We observed distinct groupings by species along PC2 such that *M. laciniatus*, the most lobed species, is distinct from *M. guttatus* and *M. nasutus*. The latter two have greater overlap along PC1 and PC2."

# PCA using covariates to explain species richness/abundance ####
data.pca <- leaf_traits_noB |>
  na.omit() |>
  select(c(20:24)) #Selecting only the log-transformed leaf trait columns. We use the log transformed data due to our findings of non-normality in the data. See this post for a hot take: https://stats.stackexchange.com/questions/164381/why-log-transforming-the-data-before-performing-principal-component-analysis

leaf_traits.pca <- as.data.frame(na.omit(leaf_traits_noB))


###Run this to create pca with prcomp function
pca <- prcomp(data.pca, scale = TRUE)
pca$rotation <- -pca$rotation
pca$x <- -pca$x

#Checking the PCA
#plot(pca,type = "lines")
#biplot(pca) # Base type PCA

# PCA using autoplot() and prcomp()and modifying with ggplot syntax ####
pca_auto <- autoplot(
  pca,
  data = leaf_traits.pca,
  alpha = 0, #Setting alpha to zero render the automatic circle point null.Manipulate shapes with geom_point().
  loadings = TRUE,
  loadings.colour = "black",
  loadings.label = TRUE,
  loadings.label.colour = "black",
  loadings.label.size = 4,
  loadings.label.vjust = 0,
  loadings.label.hjust = -0.15
) +
  geom_point(aes(fill = Species, color = Species), alpha = 0.5, size = 3) +
  geom_hline(yintercept = 0, colour = "gray45") +
  geom_vline(xintercept = 0, colour = "gray45") +
  scale_fill_manual(
    labels = c("*M. laciniatus*", "*M. nasutus*", "*M. guttatus*"),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  scale_color_manual(
    labels = c("*M. laciniatus*", "*M. nasutus*", "*M. guttatus*"),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  stat_ellipse(
    aes(color = Species),
    geom = "path",
    linewidth = 1.3,
    position = "identity",
    type = "t",
    linetype = 1,
    level = 0.95,
    segments = 51,
    na.rm = TRUE,
    show.legend = NA,
    inherit.aes = TRUE
  ) +
  #scale_x_continuous(expand = c(0.0, -0.09)) +
  theme_classic(base_size = 12) +
  theme(
    legend.text = element_markdown(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
# labs(caption = expression("ACI = anthocyanins, LPS = leaf punch strength, LMA = leaf mass per area, LT = leaf thickness. LBI = Leaf lobe index. All leaf replicates per species ("~ italic("n") ~ "= 501)."))

pca_auto$layers <- c(
  pca_auto$layers,
  pca_auto$layers[[2]],
  pca_auto$layers[[3]]
) # This adds/copies layers 2-3 and overlays them. It makes the arrows be on top of the points. There must be a better ways of doing this.

pca_auto
# layer_data(pca_auto,2)
# pca_auto + geom_point(
#   data = layer_data(pca_auto, 2),
#   aes(xend, yend),
#   size = 3
# ) +
#    geom_text(vjust=-1, label=rownames(pca_labels))

ggsave(
  filename = file.path(path, "figures/figure2.png"),
  plot = pca_auto,
  dpi = 300,
  units = c("mm"),
  width = 150,
  height = 150
)
```

## Figure 3


<!-- Seems weird to have so much Basidiomycota - how many taxa does this represent? -->

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| message: false
#| fig-width: 6
#| fig-height: 8
#| label: fig-figure3
#| fig-cap: "Relative abundance of fungal phyla in rarefied sequence data by species and elevation zone."

#Barplots with microeco
# Quick addition traits log transformed
names_list <- colnames(ps_clean_3@otu_table) #List of samples
ps_clean_3@sam_data <- plant_traits |>
  filter(Unique_ID %in% names_list) |>
  select(c(1:12, 15:20)) |>
  mutate(X = Unique_ID) |>
  column_to_rownames(var = "X") |>
  sample_data()


# Raw data
micro_psclean <- phyloseq2meco(ps_clean_3) # First convert phyloseq to meco


# t1 <- trans_abund$new(dataset = micro_psclean, taxrank = "Phylum", ntaxa = 8)
# t1gg <- t1$plot_bar(others_color = "grey70", facet = c("Species", "Elevation_cat"), xtext_keep = FALSE, legend_text_italic = FALSE, barwidth = 1, color_values = RColorBrewer::brewer.pal(8, "Set3"))
# t1gg <- t1gg +   labs(title = "Raw reads") +
#   theme(legend.title = element_text(face = "bold"),
#          strip.text = element_text(face = "italic", size = 10),
#          panel.border = element_rect(linetype = "blank", fill = NA),
#          axis.text.x = element_blank())
#t1gg

# Rarefied data
micro_rarefied_ps <- phyloseq2meco(rarefied_phyloseq) # First convert phyloseq to meco

r1 <- trans_abund$new(
  dataset = micro_rarefied_ps,
  taxrank = "Phylum",
  ntaxa = 8
)
r1gg <- r1$plot_bar(
  others_color = "grey70",
  facet = c("Species", "Elevation_cat"),
  xtext_keep = FALSE,
  legend_text_italic = FALSE,
  barwidth = 1,
  color_values = RColorBrewer::brewer.pal(8, "Set3")
)
r1gg <- r1gg +
  labs(title = "Rarefied reads") +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold.italic", size = 10),
    panel.border = element_rect(linetype = "blank", fill = NA),
    axis.text.x = element_blank()
  )
r1gg
ggsave(
  filename = file.path(path, "figures/figure3.png"),
  plot = r1gg,
  dpi = 300,
  units = c("mm"),
  width = 300,
  height = 300
)

# # Arrange plots
# ra_plots <- ggarrange(t1gg, r1gg,
#                           nrow = 2, ncol = 1,
#                           common.legend = TRUE, legend = "bottom",
#                           labels = c("A", "B"))
# ra_plots
#
# ggsave(file.path(path, "figures/figure2.png") , plot = ra_plots, dpi=300, units=c("mm"), width=200, height=250)
```



## Figure 4

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 7
#| label: fig-figure4
#| fig-cap: "Relationship between Hill orders by host species and elevation. A) Observed ASV richness (𝑞 = 0); B) Shannon's entropy (𝑞 = 1); and C) Inverse Simpson's index (𝑞 = 2) per host species as elevation increases. Blue filled points correspond to *M. laciniatus*, green filled to *M. nasutus*, and yellow to *M. guttatus*.  The solid line represents the linear model fit and the shaded area represents the 95% confidence interval."

# Shannon Diversity or hill number 1
# Three main faceted plots come out of this chunk

########## Labels ###########
#Legend title label
legend_title <- "Species"
# Species labels
labels_list <- c("M. laciniatus", "M. nasutus", "M. guttatus")
names(labels_list) <- c("M. laciniatus", "M. nasutus", "M. guttatus")
#############################

# Hill number 0
# Species richness or hill number 0

#Linear model summary
#summary(lm(Observed ~ Elevation_m, data = plant_traits_MB))
#richness_p <- format.p(cor.test(plant_traits$Observed, plant_traits$Elevation_m)$p.value) # This is just to make the p-value nicer.
# Species labels
labels_list <- c("M. laciniatus", "M. nasutus", "M. guttatus")
names(labels_list) <- c("L", "N", "G")

# Plot
hill0_diversity <- ggplot(
  data = plant_traits_MB,
  aes(x = Elevation_m, y = Observed, color = Species)
) +
  geom_jitter(
    aes(color = Species, shape = Site),
    #stat = "summary",
    #fun = mean,
    size = 3,
    alpha = 0.3
  ) +
  geom_smooth(method = lm, se = F) +
  geom_smooth(method = lm, se = T, level = 0.95, na.rm = F, color = "black") +
  stat_poly_eq(
    formula = y ~ x,
    color = "black",
    aes(
      label = paste(
        #..eq.label..,
        ..adj.rr.label.., #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    rr.digits = 2,
    p.digits = 3,
    parse = TRUE,
    label.x = "right",
    label.y = "top",
    vstep = 0.08
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    parse = TRUE,
    label.x.npc = "center",
    vstep = 0.08
  ) +
  # stat_poly_eq(formula = y ~ x, color = aes(color=Species),
  #                aes(label = paste(#..eq.label..,
  #                  ..adj.rr.label.., #"~italic(p) ==",
  #                  ..p.value.label..,
  #                  #signif(..p.value.., digits = 4),
  #                  sep = "*`,`~~")),
  #                rr.digits = 2,
  #                p.digits = 3 ,
  #                parse = TRUE,
  #                label.x = "left",
  #                label.y= "top",
  #                vstep = 0.08) +
  scale_fill_manual(
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  scale_color_manual(
    legend_title,
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  scale_shape_manual(values = c(1:22)) +
  theme_lfdiv_continuous +
  labs(y = "Alpha diversity (\U1D492 = 0)", x = "", col = "") +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.05, 0.05))
#facet_wrap(~ Species) +
#guides(color = "none")        #To turn off color legend
# annotate(geom = "text", x = 2100, y = 2.5, label = "p = 0.03", color = "black")
#annotate(geom="text", x=2100, y=50, label="p = 0.06", color = "black")

#hill0_diversity

#ggsave(filename = file.path(path, "figures/richness_elevation.png"), plot = hill0_diversity, dpi=300, units=c("mm"), width=250, height=150)

# Linear models per species
# Linear model summary
# M. laciniatus
#summary(lm(Observed ~ Elevation_m, data = plant_traits[plant_traits$Species == "L",]))
#Not significant

# M. nasutus
#summary(lm(Observed ~ Elevation_m, data = plant_traits[plant_traits$Species == "N",]))
#Not significant

# M. guttatus
#summary(lm(Observed~ Elevation_m, data = plant_traits[plant_traits$Species == "G",]))
#Not significant

# Hill number 1
# Shannon diversity or hill number 1
# Linear model summary
#summary(lm(hill_1 ~ Elevation_m, data = plant_traits_MB))
#elev_div_p <- format.p(cor.test(plant_traits$Shannon, plant_traits$Elevation_m)$p.value) # This is just to make the p-value nicer.

# Plot
hill1_diversity <- ggplot(
  data = plant_traits_MB,
  aes(x = Elevation_m, y = hill_2, color = Species)
) +
  geom_jitter(
    aes(color = Species, shape = Site),
    #stat = "summary",
    #fun = mean,
    size = 3,
    alpha = 0.3
  ) +
  geom_smooth(method = lm, se = F) +
  geom_smooth(method = lm, se = T, level = 0.95, na.rm = F, color = "black") +
  stat_poly_eq(
    formula = y ~ x,
    color = "black",
    aes(
      label = paste(
        #..eq.label..,
        ..adj.rr.label.., #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    rr.digits = 2,
    p.digits = 3,
    parse = TRUE,
    label.x = "right",
    label.y = "top",
    vstep = 0.08
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    parse = TRUE,
    label.x.npc = "center",
    vstep = 0.08
  ) +
  scale_color_manual(
    legend_title,
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  scale_shape_manual(values = c(1:22)) +
  labs(y = "Alpha diversity (\U1D492 = 1)", x = "", col = "") +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  theme_lfdiv_continuous +
  #theme(axis.title.x = element_text()) +
  theme(strip.text = element_blank())
#facet_wrap(~ Species) +
#guides(color = "none")        #To turn off color legend

#hill1_diversity

#ggsave(filename = file.path(path, "figures/shannon_elevation.png"), plot = hill1_diversity, dpi=300, units=c("mm"), width=250, height=150)

# Linear models per species
# Linear model summary
# M. laciniatus
#summary(lm(hill_1 ~ Elevation_m, data = plant_traits[plant_traits$Species == "L",]))
#Not significant

# M. nasutus
#summary(lm(hill_1 ~ Elevation_m, data = plant_traits[plant_traits$Species == "N",]))
#Not significant

# M. guttatus
#summary(lm(hill_1 ~ Elevation_m, data = plant_traits[plant_traits$Species == "G",]))
#Not significant

# Hill number 2
# Inverse Simpson diversity or hill number 2
# Linear model summary
#summary(lm(hill_2 ~ Elevation_m, data = plant_traits))
#elev_div_p <- format.p(cor.test(plant_traits$Shannon, plant_traits$Elevation_m)$p.value) # This is just to make the p-value nicer.

# Plot
hill2_diversity <- ggplot(
  data = plant_traits_MB,
  aes(x = Elevation_m, y = hill_2, color = Species)
) +
  geom_jitter(
    aes(color = Species, shape = Site),
    #stat = "summary",
    #fun = mean,
    size = 3,
    alpha = 0.3
  ) +
  geom_smooth(method = lm, se = F) +
  geom_smooth(method = lm, se = T, level = 0.95, na.rm = F, color = "black") +
  stat_poly_eq(
    formula = y ~ x,
    color = "black",
    aes(
      label = paste(
        #..eq.label..,
        ..adj.rr.label.., #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    rr.digits = 2,
    p.digits = 3,
    parse = TRUE,
    label.x = "right",
    label.y = "top",
    vstep = 0.08
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    parse = TRUE,
    label.x.npc = "center",
    vstep = 0.08
  ) +
  #scale_fill_manual(labels = c("*M. laciniatus*","*M. nasutus*", "*M. guttatus*"),
  #values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))) +
  scale_color_manual(
    legend_title,
    labels = c("M. laciniatus", "M. nasutus", "M. guttatus"),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  scale_shape_manual(values = c(1:22)) +
  labs(
    y = "Alpha diversity (\U1D492 = 2)",
    x = "Elevation (m.a.s.l.)",
    col = ""
  ) +
  theme_lfdiv_continuous +
  theme(strip.text = element_blank())
# facet_wrap(~ Species)
#guides(color = "none")        #To turn off color legend

#hill2_diversity

#ggsave(filename = file.path(path, "figures/invsimpson_elevation.png"), plot = hill2_diversity, dpi=300, units=c("mm"), width=250, height=150)

# Linear models per species
# Linear model summary
# M. laciniatus
#summary(lm(hill_2 ~ Elevation_m, data = plant_traits[plant_traits$Species == "L",]))
#Not significant

# M. nasutus
#summary(lm(hill_2 ~ Elevation_m, data = plant_traits[plant_traits$Species == "N",]))
#Not significant

# M. guttatus
#summary(lm(hill_2 ~ Elevation_m, data = plant_traits[plant_traits$Species == "G",]))
#Not significant

###### Joined plots ######
hill_numbers <-
  ggarrange(
    hill0_diversity,
    hill1_diversity,
    hill2_diversity,
    common.legend = TRUE,
    legend = "right",
    labels = c("A", "B", "C"),
    nrow = 3,
    ncol = 1
  )

hill_numbers
ggsave(
  filename = file.path(path, "figures/figure4.png"),
  plot = hill_numbers,
  dpi = 300,
  units = c("mm"),
  width = 275,
  height = 300
)
```



## Figure 5

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 5
#| fig-height: 6
#| label: fig-figure5
#| fig-cap: "FEF community composition association with leaf functional traits and elevation by host species. Distance-based Redundancy Analysis (dbRDA) plot of rarefied FEF community and leaf functional traits by species (colors) and by elevation zone (shapes). Each cluster of points represents a rarefied FEF community sample from one host species. Solid arrow lines represent significant associations (*p* < .01)with non-significant leaf trait associations left out for clarity. The length and direction of the arrows indicate the strength and direction of the association between the traits and the FEF community composition. Ellipses represent 95% confidence intervals. The plot is based on the Bray-Curtis dissimilarity matrix."

# dbRDA using Elevation as continuous variable
#logACI + logLT + logLPS + logLMA +
m5_hell <- readRDS("data/clean_data/statistics/dbrda_m5_hell.rds")

# Final model
final_model <- m5_hell # Previously m4_hell was the best

B <- summary(final_model)

# Extracting scores for the plot
ordination_scores <- as.data.frame(vegan::scores(final_model)$sites)

# Extracting names for the plot
ordination_names <- rownames(ordination_scores)

# Row names to column
ordination_scores$Unique_ID <- rownames(ordination_scores)

# Subset for the plot
dbrda_df <- plant_traits_MB
#filter(Unique_ID %in% ordination_names)

# Merge scores with trait data
merged_data <- right_join(
  ordination_scores,
  dbrda_df,
  by = c("Unique_ID" = "X")
)

# Extracting scores for arrows
arrow_scores <- as.data.frame(vegan::scores(final_model, display = "bp"))

# Define the arrow aesthetic mapping
arrow_map <- aes(
  xend = (dbRDA1 * 2),
  yend = (dbRDA2 * 2),
  x = 0,
  y = 0,
  shape = NULL,
  color = NULL,
  fill = NULL
)
label_map <- aes(
  x = 2 * dbRDA1,
  y = 2 * dbRDA2,
  label = row.names(arrow_scores),
  shape = NULL,
  color = NULL,
  fill = NULL
)
arrowhead <- arrow(length = unit(0.02, "npc"))

#subset A4 for labeling
arrow_scores <- arrow_scores[order(rownames(arrow_scores)), ]
A4 <- lapply(1:6, function(i) {
  arrow_scores[i, ]
})

# dbRDA plot based on species
p_final_model <-
  ggplot(data = merged_data, aes(x = dbRDA1, y = dbRDA2)) +
  geom_point(
    data = merged_data,
    aes(color = Species, shape = Elevation_cat),
    size = rel(4),
    alpha = 0.2,
    stroke = 1
  ) +
  stat_ellipse(
    aes(color = Species),
    geom = "path",
    linewidth = 1.3,
    position = "identity",
    type = "t",
    linetype = 1,
    level = 0.95,
    segments = 51,
    na.rm = FALSE,
    show.legend = NA,
    inherit.aes = TRUE
  ) +
  geom_segment(
    arrow_map,
    linewidth = rel(.8),
    data = arrow_scores[1:2, ],
    color = "black",
    arrow = arrowhead
  ) +
  geom_vline(xintercept = 0, color = "grey30", linetype = "dashed") + # plot vertical line
  geom_hline(yintercept = 0, color = "grey30", linetype = "dashed") + # plot horizontal line
  geom_text(
    label_map,
    size = rel(4),
    data = A4[[1]],
    fontface = "bold",
    show.legend = FALSE,
    label = c("Elev."),
    nudge_x = 0.45
  ) +
  geom_text(
    label_map,
    size = rel(4),
    data = A4[[2]],
    fontface = "bold",
    show.legend = FALSE,
    label = c("logLBI"),
    nudge_y = 0.2
  ) +
  xlab(
    label = paste(
      "dbRDA1 (",
      round(B$concont$importance[2, 1] * 100, digits = 1),
      "%)",
      sep = ""
    )
  ) +
  ylab(
    label = paste(
      "dbRDA2 (",
      round(B$concont$importance[2, 2] * 100, digits = 1),
      "%)",
      sep = ""
    )
  ) +
  scale_color_manual(
    labels = c("*M. laciniatus*", "*M. nasutus*", "*M. guttatus*"),
    values = met.brewer(
      "Isfahan2",
      n = 3,
      type = "discrete",
      direction = -1
    )
  ) +
  scale_shape_manual(values = c(1:3)) +
  theme_classic(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    legend.text = element_markdown(),
    legend.position = "bottom",
    plot.caption = element_markdown(),
    strip.text = element_text(face = "bold")
  ) +
  labs(caption = "PERMDISP: *F*<sub>7,4192</sub> = 268.1, *p* = .001") #Results from beta_perm3
#facet_wrap(~ Elevation_cat)

# Rearranging the order of layers so arrows are above the points.
#p_final_model$layers
# p_final_model$layers <- c(p_final_model$layers, p_final_model$layers[[2]], p_final_model$layers[[5]], p_final_model$layers[[6]])
p_final_model

#ggsave(filename=file.path(path, "figures/dbRDA_elevation_labels.svg"), plot = p_final_model, dpi=300, units=c("mm"), width=250, height=150)

#Save the plot
ggsave(
  file.path(path, "figures/figure5.png"),
  plot = p_final_model,
  dpi = 300,
  units = c("mm"),
  width = 225,
  height = 225
)
```

## Figure 6

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 7
#| label: fig-figure6
#| fig-cap: "Relationship between β-diversity and elevation by host species. The solid line represents simple linear regression and the shaded area represents the 95% confidence interval."

# Mean-Bray-Curtis dissimilarity
mean_bray <- vegdist(dbrda_hell_matrix, method = "bray") |>
  as.matrix() |>
  as.data.frame()

mean_bray <- rownames_to_column(mean_bray, var = "Unique_ID") |>
  mutate(mean_bray = rowMeans(mean_bray)) |>
  select(Unique_ID, mean_bray)

# Joining to dbrda_traits to avoid duplicate DFs
dbrda_traits_imp <- left_join(
  dbrda_traits_imp,
  mean_bray,
  by = c("X" = "Unique_ID")
)

# Simple linear regressions for the stats
# summary(lm(mean_bray ~ Elevation_m, data =subset(dbrda_traits_imp, Species == "M. laciniatus")))
# summary(lm(mean_bray ~ Elevation_m, data =subset(dbrda_traits_imp, Species == "M. nasutus")))
# summary(lm(mean_bray ~ Elevation_m, data =subset(dbrda_traits_imp, Species == "M. guttatus")))

# GGplot
beta_elevation <- ggplot(
  data = dbrda_traits_imp,
  aes(x = Elevation_m, y = mean_bray, color = Species)
) +
  geom_point(
    aes(color = Species),
    size = rel(4),
    alpha = 0.2,
    stroke = 1
  ) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Elevation (m)", y = "β-diversity") +
  scale_color_manual(
    labels = c("*M. laciniatus*", "*M. nasutus*", "*M. guttatus*"),
    values = met.brewer(
      "Isfahan2",
      n = 3,
      type = "discrete",
      direction = -1
    )
  ) +
  stat_poly_eq(
    formula = y ~ x,
    color = "black",
    aes(
      label = paste(
        #..eq.label..,
        after_stat(adj.rr.label),
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    #method = "loess",
    rr.digits = 2,
    p.digits = 3,
    parse = TRUE,
    label.x = "right",
    label.y = "top",
    vstep = 0.08
  ) +
  theme_classic(base_size = 16) +
  theme(
    legend.title = element_text(face = "bold"),
    legend.text = element_markdown(),
    legend.position = "bottom",
    plot.caption = element_markdown(),
    strip.text = element_text(face = "bold.italic", size = 14)
  ) +
  facet_wrap(~Species)

beta_elevation

ggsave(
  file.path(path, "figures/figure6.png"),
  plot = beta_elevation,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 250
)
```


## Figure 7

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 7
#| label: fig-figure7
#| fig-cap: "Dissimilarity of FEF communities correlates with *Mimulus* as a function of geographic distance per site. Data represent pairwise Bray-Curtis dissimilarity (β-diversity) among sites. A) FEF community dissimilarity in considering all host species. B) FEF community dissimilarity by host species. Blue filled points correspond to *M. laciniatus*, green filled to *M. nasutus*, and yellow to *M. guttatus*."

# Prepare the data for correlation plot
# Longer format
asv_vegdist <- as.matrix(vegdist(t(asv_matrix), method = "bray"))

asv_long <- asv_vegdist |>
  as_tibble(rownames = "A") |>
  pivot_longer(-A, names_to = "B", values_to = "bray_diss")

geo_long <- geo_distm |>
  as_tibble(rownames = "A") |>
  pivot_longer(-A, names_to = "B", values_to = "geo_dist")

# Merge the two dataframes
asv_geo_matrix_long <-
  inner_join(asv_long, geo_long, by = c("A", "B")) |>
  inner_join(
    plant_traits |>
      dplyr::select(Unique_ID, Site, Species, Elevation_cat),
    by = c("A" = "Unique_ID")
  )

# Mantel test
# All data points
set.seed(859)
abund_geo <- mantel(
  asv_vegdist,
  geo_distm,
  method = "spearman",
  parallel = 8,
  permutations = 999,
  na.rm = TRUE
)
#abund_geo

# Mantel test by species
# M. laciniatus

# Samples with species to subset/filter matrices downstream
# M. laciniatus
lac_names_rfy <- plant_traits |>
  slice(1:157) |> # Only selecting observations with ASV data
  filter(Unique_ID %in% final_names_methodB, Species == "M. laciniatus") |>
  pull(Unique_ID)

# Subset the data
lac_mantel <- mantel(
  asv_long |> # Subset the data
    filter(A %in% lac_names_rfy, B %in% lac_names_rfy) |>
    pivot_wider(names_from = B, values_from = bray_diss) |>
    column_to_rownames(var = "A") |>
    as.matrix(),
  geo_long |> # Subset the data
    filter(A %in% lac_names_rfy, B %in% lac_names_rfy) |>
    pivot_wider(names_from = B, values_from = geo_dist) |>
    column_to_rownames(var = "A") |>
    as.matrix(),
  method = "spearman",
  permutations = 999,
  na.rm = TRUE
)
#lac_mantel

# M. nasutus
nas_names_rfy <- plant_traits |>
  slice(1:157) |> # Only selecting observations with ASV data
  filter(Unique_ID %in% final_names_methodB, Species == "M. nasutus") |>
  pull(Unique_ID)
nas_mantel <- mantel(
  asv_long |> # Subset the data
    filter(A %in% nas_names_rfy, B %in% nas_names_rfy) |>
    pivot_wider(names_from = B, values_from = bray_diss) |>
    column_to_rownames(var = "A") |>
    as.matrix(),
  geo_long |> # Subset the data
    filter(A %in% nas_names_rfy, B %in% nas_names_rfy) |>
    pivot_wider(names_from = B, values_from = geo_dist) |>
    column_to_rownames(var = "A") |>
    as.matrix(),
  method = "spearman",
  permutations = 999,
  na.rm = TRUE
)
#nas_mantel

# M. guttatus
gut_names_rfy <- plant_traits |>
  slice(1:157) |> # Only selecting observations with ASV data
  filter(Unique_ID %in% final_names_methodB, Species == "M. guttatus") |>
  pull(Unique_ID)
# Subset the data
gut_mantel <- mantel(
  asv_long |> # Subset the data
    filter(A %in% gut_names_rfy, B %in% gut_names_rfy) |>
    pivot_wider(names_from = B, values_from = bray_diss) |>
    column_to_rownames(var = "A") |>
    as.matrix(),
  geo_long |> # Subset the data
    filter(A %in% gut_names_rfy, B %in% gut_names_rfy) |>
    pivot_wider(names_from = B, values_from = geo_dist) |>
    column_to_rownames(var = "A") |>
    as.matrix(),
  method = "spearman",
  permutations = 999,
  na.rm = TRUE
)
#gut_mantel

# Visualize the correlation
# Species labels
labels_list <- c("M. laciniatus", "M. nasutus", "M. guttatus")
names(labels_list) <-
  c("M. laciniatus", "M. nasutus", "M. guttatus")

# Correlation ggplot
mantel_test <-
  ggplot(asv_geo_matrix_long, aes(y = bray_diss, x = geo_dist)) +
  geom_point(aes(color = Species), alpha = 0.5, size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(
    values = met.brewer(
      "Isfahan2",
      n = 4,
      type = "discrete",
      direction = c(-1)
    )
  ) +
  labs(y = "β-diversity", x = "Geographic distance (m)", col = "") +
  scale_x_continuous(
    labels = function(x) {
      format(x, scientific = FALSE)
    }
  ) +
  theme_lfspp_continuous

# Adding the mantel test results
italic_r <- expression(italic("r ="))
italic_p <- expression(italic("p ="))
statistic <- round(abund_geo$statistic, digits = 2)
signif <- round(abund_geo$signif, digits = 3)


mantel_test <- mantel_test +
  ggplot2::annotate(
    "text",
    x = 110000,
    y = 0,
    label = paste(italic_r, statistic, italic_p, signif, sep = "~"),
    size = 5,
    parse = TRUE
  )
#mantel_test

# By species
mantel_species <-
  ggplot(asv_geo_matrix_long, aes(y = bray_diss, x = geo_dist)) +
  geom_jitter(aes(color = Species), alpha = 0.5, size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_manual(
    legend_title,
    values = met.brewer(
      "Isfahan2",
      n = 4,
      type = "discrete",
      direction = c(-1)
    )
  ) +
  labs(y = "β-diversity", x = "Geographic distance (m)", col = "") +
  scale_x_continuous(
    labels = function(x) {
      format(x, scientific = FALSE)
    }
  ) +
  theme_lfspp_continuous +
  theme_classic(base_size = 16) +
  theme(
    legend.title = element_text(face = "bold"),
    panel.border = element_rect(linetype = "blank", fill = NA),
    legend.text = element_text(face = "italic"),
    strip.text = element_text(face = "bold.italic", size = 14),
    axis.text.x = element_text(size = 10),
    legend.position = "bottom"
  ) +
  guides(guide_legend(color = "Species")) +
  facet_wrap(~Species)
#mantel_species

# Adding the mantel test results

italic_r <- expression(italic("r ="))
italic_p <- expression(italic("p ="))
statistic_lac <- round(lac_mantel$statistic, digits = 2)
signif_lac <- round(lac_mantel$signif, digits = 3)
statistic_nas <- round(nas_mantel$statistic, digits = 2)
signif_nas <- round(nas_mantel$signif, digits = 3)
statistic_gut <- round(gut_mantel$statistic, digits = 3)
signif_gut <- round(gut_mantel$signif, digits = 3)

my_tags <-
  c(
    paste0("r = ", statistic_lac, ", p = ", signif_lac, sep = " "),
    paste0("r = ", statistic_nas, ", p = ", signif_nas, sep = " "),
    paste0("r = ", statistic_gut, ", p = ", signif_gut, sep = " ")
  )

mantel_species <- tag_fat(
  mantel_species,
  x = -Inf,
  y = -Inf,
  vjust = -1,
  hjust = -0.95,
  open = "",
  close = "",
  fontface = 3,
  size = 4,
  tag_pool = my_tags
)


# Joining plots
# mantel <- ggpubr::ggarrange(mantel_test, "", mantel_species,
#                     ncol = 1, nrow = 3,
#                     widths = c(1, 0.1, 1),
#                     heights = c(1, 0.1, 1),
#                     labels = c("A", "B"),
#                     common.legend = T, legend = "bottom")

mantel_species
ggsave(
  file.path(path, "figures/figure7.png"),
  plot = mantel_species,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 200
)
```


## Table 1

```{r, table1}
#| echo: false
#| eval: true
#| tidy: true
#| label: tbl-table1
#| tbl-cap: "Generalized linear mixed models (GLMMs) for predicting mean Bray-Curtis dissimilarity"

load("data/clean_data/statistics/glmm_models.rda")

# Huxtable to flextable
table1 <- huxreg(
  "Model 1" = B1REM,
  "Model 2" = B3REM,
  "Model 3" = B4REM,
  "Model 4" = B5REM,
  coefs = c(
    "Intercept" = "(Intercept)",
    "M. nasutus" = "SpeciesM. nasutus",
    "M. guttatus" = "SpeciesM. guttatus",
    "Elevation" = "Elevation_m",
    "ACI" = "logACI",
    "LT" = "logLT",
    "LPS" = "logLPS",
    "LMA" = "logLMA",
    "LBI" = "logLBI"
  ),
  error_pos = "right",
  error_format = "({conf.low} - {conf.high} ) [{statistic}]",
  number_format = "%.2f",
  align = ".",
  ci_level = 0.95,
  stars = c(`***` = 0.001, `**` = 0.01, `*` = 0.05),
  bold_signif = 0.05,
  borders = 0.4,
  outer_borders = 0.8,
  note = "Significance levels are represented by asterisks [p < 0.05 (*), p < 0.01 (**), p < 0.001 (***), and p < 0.0001 (****)]. T statistics in brackets.",
  statistics = c("N" = "nobs", "logLik", "AIC")
) |>
  insert_row(
    "Predictors",
    "Estimate",
    "CI [t-statistic]",
    "Estimate",
    "CI [t-statistic]",
    "Estimate",
    "CI [t-statistic]",
    "Estimate",
    "CI [t-statistic]",
    after = 1
  ) |>
  insert_row("Random effects", "", "", "", "", "", "", "", "", after = 11) |>
  set_all_padding(10) |>
  set_outer_padding(0) |>
  set_bold(row = 1, col = everywhere) |>
  #set_bottom_border(row = 1, col = everywhere) |>
  set_width(1.1) |>
  theme_article() |> #flextable::theme_apa() is good too.
  huxtable::as_flextable(column_to_header = TRUE) |>
  hline(
    12,
    part = "all",
    border = fp_border(
      color = "black",
      style = "solid",
      width = 1
    )
  ) |>
  hline(
    15,
    part = "all",
    border = fp_border(
      color = "black",
      style = "solid",
      width = 1.25
    )
  ) |>
  hline(
    i = 1,
    j = c(2:9),
    part = "body",
    border = fp_border(
      color = "black",
      style = "solid",
      width = 1
    )
  ) |>
  flextable::align(
    i = 1:14,
    j = 2:9,
    align = "center",
    part = "body"
  ) |>
  flextable::style(i = c(1:16), j = 1, pr_t = fp_text(bold = FALSE)) |>
  flextable::italic(
    i = c(4, 5, 12, 15),
    j = 1,
    italic = TRUE,
    part = "body"
  ) |>
  flextable::bold(
    i = c(2, 12),
    j = 1,
    bold = TRUE,
    part = "body"
  ) |>
  flextable::set_caption(
    "Generalized linear mixed models (GLMMs) for predicting mean Bray-Curtis dissimilarity"
  ) |>
  flextable::font(
    i = NULL,
    j = NULL,
    fontname = "Tex Gyre Termes",
    part = "all",
    cs.family = fontname,
    hansi.family = fontname,
    eastasia.family = fontname
  )
# add_footer_lines(
#   value = as_paragraph("Significance levels are represented by asterisks [*p* < .05 (\\*), *p* < .01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]. T statistics in brackets.")
# )
table1

#save_as_docx(table1, path = "tables/model_comparison.docx")
#save_as_image(table1, path = "tables/model_comparison.svg")
```


# Supplementary Material

## Figure S1

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| message: false
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM21
#| fig-cap: "Figure S1. Correlation matrix of log-transformed leaf functional traits by host species. The plot is based on the Spearman's rho. Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < 0.01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

t4 <- trans_env$new(dataset = micro_psclean, env_cols = c(5, 14:18))

t4gg <- t4$cal_autocor(
  group = "Species",
  upper = list(continuous = GGally::wrap("cor", method = "spearman"))
) +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 10),
    panel.border = element_rect(linetype = "blank", fill = NA)
  ) +
  scale_fill_manual(
    values = met.brewer("Isfahan2", n = 3, type = "discrete", direction = c(-1))
  ) +
  scale_color_manual(
    values = met.brewer("Isfahan2", n = 3, type = "discrete", direction = c(-1))
  )

t4gg
ggplot2::ggsave(
  file.path(path, "figures/supp_fig1.jpg"),
  plot = t4gg,
  dpi = 300,
  units = c("mm"),
  width = 300,
  height = 300
)
```

## Figure S2

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| message: false
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM22
#| fig-cap: "Figure S2. Heatmap of the top 10 most abundant FEF genera in the dataset. The heatmap is faceted by host species and elevation zone."

# Plot with micro_psclean
t2 <-
  trans_abund$new(dataset = micro_psclean, taxrank = "Genus", ntaxa = 10)
t2gg <-
  t2$plot_heatmap(
    facet = c("Species", "Elevation_cat"),
    xtext_keep = FALSE,
    withmargin = FALSE
  ) +
  labs(title = "Un-rarefied reads") +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold.italic", size = 12),
    panel.border = element_rect(linetype = "blank", fill = NA),
    axis.text.x = element_blank()
  )

#t2gg

r2 <-
  trans_abund$new(dataset = micro_rarefied_ps, taxrank = "Genus", ntaxa = 10)
r2gg <-
  r2$plot_heatmap(
    facet = c("Species", "Elevation_cat"),
    xtext_keep = FALSE,
    withmargin = FALSE
  ) +
  labs(title = "Rarefied reads") +
  theme(
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold.italic", size = 12),
    panel.border = element_rect(linetype = "blank", fill = NA),
    axis.text.x = element_blank()
  )

r2gg
ggsave(
  filename = file.path(path, "figures/supp_fig2b.jpg"),
  plot = r2gg,
  dpi = 300,
  units = c("mm"),
  width = 300,
  height = 250
)

# Joined heatplots
heatplots <- ggarrange(
  t2gg,
  r2gg,
  ncol = 1,
  nrow = 2,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A", "B"),
  heights = c(1, 1)
)
#heatplots

# Save the plot
ggplot2::ggsave(
  filename = file.path(path, "figures/supp_fig2.png"),
  plot = heatplots,
  dpi = 300,
  units = c("mm"),
  width = 300,
  height = 250
)
```


## Figure S3
```{r}
#| echo: false
#| eval: true
#| tidy: true
#| message: false
#| fig-height: 6
#| fig-width: 6
#| label: suppfigMIM23
#| fig-cap: "Figure S3. Venn diagrams of presence and absence of ASVs among species and elevation zones un-rarified data and rarified data. A) Represents the overlap of ASVs in in host species and B) per elevation zone. Number of ASVs are represented the count and in parentheses as a ratio shared between host species and elevation zone"

# un-rarified data
# Per Species
dataset1 <- micro_psclean$merge_samples(use_group = "Species")
# create trans_venn object
v1 <-
  trans_venn$new(dataset1, ratio = "numratio") # Percentage represent ratio of shared ASVs
v1gg <-
  v1$plot_venn() + #https://r-charts.com/color-palettes/#discrete
  ggtitle("Un-rarified data") +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16,
      hjust = 0.5,
      vjust = 0.5
    )
  )


# Per Elevations
dataset2 <- micro_psclean$merge_samples(use_group = "Elevation_cat")
v2 <- trans_venn$new(dataset2, ratio = "numratio")
#v2$plot_venn()

# rarified data
#micro_rarefied_ps <- phyloseq2meco(rarefied_phyloseq) # First convert phyloseq to meco
# Per Species
dataset3 <- micro_rarefied_ps$merge_samples(use_group = "Species")
# create trans_venn object
v3 <- trans_venn$new(dataset3, ratio = "numratio")
v3gg <- v3$plot_venn() +
  #ggtitle("Rarified data") +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16,
      hjust = 0.5,
      vjust = 0.5
    )
  )

# Per Elevations
dataset4 <- micro_rarefied_ps$merge_samples(use_group = "Elevation_cat")
v4 <- trans_venn$new(dataset4, ratio = "numratio")
#v4$plot_venn()

#
# # joined plots
# venn_plots <-
#   ggarrange(
#     v1gg,
#     v3gg,
#     "", # Fake plot to create space
#     "",
#     v2$plot_venn(),
#     v4$plot_venn(),
#     ncol = 2,
#     nrow = 3,
#     common.legend = TRUE,
#     heights = c(1, 0.1, 1),
#     legend = "bottom",
#     labels = c("A", "B","", "", "C", "D")
#   )

# venn_plots

rrfy_venn <-
  ggarrange(
    v3gg,
    v4$plot_venn(),
    ncol = 2,
    nrow = 1,
    common.legend = TRUE,
    legend = "bottom",
    labels = c("A", "B")
  ) +
  ggtitle("Rarified data")

rrfy_venn

ggsave(
  filename = file.path(path, "figures/supp_fig3.jpg"),
  plot = rrfy_venn,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 225
)

# # Save
# ggsave(filename = file.path(path, "figures/supp_fig3.png"), plot = venn_plots, dpi=300, units=c("mm"), width=300, height=300)
```

## Figure S4

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM24
#| fig-cap: "Figure S4. Alpha and beta diversity means comparisons in host species and elevation zones. Alpha diversity means comparisons; A) Observed ASV richness (𝑞 = 0); B) Shannon's entropy (𝑞 = 1); and C) Inverse Simpson's index (𝑞 = 2) per host species. A-C) Blue filled boxplots correspond to *M. laciniatus*, green filled to *M. nasutus*, and yellow to *M. guttatus*. Beta diversity mean comparisons; A) Observed ASV richness (𝑞 = 0); B) Shannon's entropy (𝑞 = 1); and C) Inverse Simpson's index (𝑞 = 2) per elevation zone. D-F) Violet boxplots correspond to LOW elevation sites, pink filled to MID and light maroon to HIGH elevation sites, while squares represent *M. laciniatus*, circles *M. nasutus* and triangles *M. guttatus*.  Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < .01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

############################################
#### Alpha diversity ####
#### At the species level ####
############################################

labels_list <- c("M. laciniatus", "M. nasutus", "M. guttatus")
names(labels_list) <- c("M. laciniatus", "M. nasutus", "M. guttatus")

# Plot
hill0_species <- ggplot(
  data = plant_traits_MB,
  aes(x = Species, y = Observed)
) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(
    aes(color = Species),
    alpha = 0.5,
    outlier.alpha = 0.0
  ) +
  scale_color_manual(
    name = "Species",
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = -1)
  ) +
  theme_sppdiv_discrete +
  labs(y = "Alpha diversity (\U1D492 = 0)", x = "", col = "") +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  )
#guides(color = "none")        #To turn off color legend

#hill0_species
# annotate(geom = "text", x = 2100, y = 2.5, label = "p = 0.03", color = "black")
#annotate(geom="text", x=2100, y=50, label="p = 0.06", color = "black")

# Hill number 1
# Shannon diversity or hill number 1
# Plot
hill1_species <- ggplot(data = plant_traits_MB, aes(x = Species, y = hill_1)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(
    aes(color = Species),
    alpha = 0.5,
    outlier.alpha = 0.0
  ) +
  scale_color_manual(
    name = "Species",
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = -1)
  ) +
  #scale_color_manual(aes(color = "black")) +
  theme_sppdiv_discrete +
  theme(strip.text = element_text(face = "bold", size = 14)) +
  #labs(y =  "Alpha and Beta diversity (\U1D492 = 1)", x = "", col = "") +
  labs(y = "Alpha diversity (\U1D492 = 1)", x = "", col = "") +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  )
#facet_wrap(~ Elevation_cat)
# guides(color = "none")        #To turn off color legend

#hill1_species

# Hill number 2
# Inverse Simpson diversity or hill number 2
hill2_species <- ggplot(data = plant_traits_MB, aes(x = Species, y = hill_2)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(
    aes(color = Species),
    alpha = 0.5,
    outlier.alpha = 0.0
  ) +
  scale_color_manual(
    name = "Species",
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = -1)
  ) +
  theme_sppdiv_discrete +
  labs(y = "Alpha diversity (\U1D492 = 2)", x = "", col = "") +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  )
# facet_wrap(~ Species, labeller = labeller(Species = labels_list)) +
#guides(color = "none")        #To turn off color legend

#hill2_species

# Joined plots
hill_numbers_species <- ggarrange(
  hill0_species,
  hill1_species,
  hill2_species,
  nrow = 1,
  ncol = 3,
  common.legend = TRUE,
  labels = c("A", "B", "C"),
  legend = "bottom"
)

#hill_numbers_species
#ggsave(filename = file.path(path, "figures/hill_numbers_species.png"), plot = hill_numbers_species, dpi=300, units=c("mm"), width=250, height=275)

#################################
#### Beta diversity ####
#### Across elevation ####
#################################

# Observed richness
# Hill number 0
hill0_beta <-
  ggplot(plant_traits_MB, aes(x = Elevation_cat, y = Observed)) +
  geom_point(
    aes(color = Elevation_cat),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Elevation_cat), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    name = "Elevation",
    labels = c('Low', 'Mid', 'High'),
    values = met.brewer(
      "Cassatt1",
      n = 3,
      type = "discrete",
      direction = -1
    )
  ) +
  scale_shape_manual(values = c(15:18)) +
  theme_elevdiv_discrete +
  theme(legend.text = element_text(size = 14)) +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  stat_compare_means(
    comparisons = elevation_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  labs(y = "Beta diversity (\U1D492 = 0)", x = "") +
  guides(
    fill = guide_legend(title = "Elevation"),
    shape = guide_legend(title = "Species")
  )

#hill0_beta

#ggsave(file.path(path, "figures/shannon_elevation.png") , plot = bbe, dpi=300, units=c("mm"), width=200, height=150)

# Hill number 1
# Shannon diversity or hill number 1
hill1_beta <-
  ggplot(plant_traits_MB, aes(x = Elevation_cat, y = hill_1)) +
  geom_point(
    aes(color = Elevation_cat),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Elevation_cat), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    name = "Elevation",
    labels = c('Low', 'Mid', 'High'),
    values = met.brewer(
      "Cassatt1",
      n = 3,
      type = "discrete",
      direction = -1
    )
  ) +
  scale_shape_manual(values = c(15:18)) +
  theme_elevdiv_discrete +
  theme(legend.text = element_text(size = 14)) +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  stat_compare_means(
    comparisons = elevation_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  labs(y = "Beta diversity (\U1D492 = 1)", x = "") +
  guides(
    fill = guide_legend(title = "Elevation"),
    shape = guide_legend(title = "Species")
  )

#hill1_beta

# Hill number 2
# Inverse Simpson diversity or hill number 2

hill2_beta <- ggplot(plant_traits_MB, aes(x = Elevation_cat, y = hill_2)) +

  geom_point(
    aes(color = Elevation_cat),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Elevation_cat), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    name = "Elevation",
    labels = c('Low', 'Mid', 'High'),
    values = met.brewer(
      "Cassatt1",
      n = 3,
      type = "discrete",
      direction = -1
    )
  ) +
  scale_shape_manual(values = c(15:18)) +
  theme_elevdiv_discrete +
  theme(legend.text = element_text(size = 14)) +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  stat_compare_means(
    comparisons = elevation_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  labs(y = "Beta diversity (\U1D492 = 2)", x = "") +
  guides(
    fill = guide_legend(title = "Elevation"),
    shape = guide_legend(title = "Species")
  )

#hill2_beta

# Joining plots
hill_beta <- ggarrange(
  hill0_beta,
  hill1_beta,
  hill2_beta,
  nrow = 1,
  ncol = 3,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("D", "E", "F")
)
# hill_beta

#### Alpha and Beta diversity ####
# Joined plots
# A joined plot of joined plots

alpha_beta_diversity <- ggarrange(
  hill_numbers_species,
  "",
  hill_beta,
  nrow = 3,
  ncol = 1,
  common.legend = FALSE,
  widths = c(1, 0.1, 1),
  heights = c(1, 0.1, 1),
  legend = "bottom"
)

alpha_beta_diversity

ggsave(
  filename = file.path(path, "figures/supp_fig4.jpg"),
  plot = alpha_beta_diversity,
  dpi = 300,
  units = c("mm"),
  width = 300,
  height = 275
)

# Alpha and beta diversity plot for defense talk
# defense_alpha_beta <- ggarrange(hill1_diversity, hill1_species,
#                           nrow = 2, ncol = 1,
#                           common.legend = TRUE,
#                           legend = "bottom",
#                           labels = c("A", "B"))
# defense_alpha_beta
#
# ggsave(filename = file.path(path, "figures/defense_alpha_beta.svg"), plot = defense_alpha_beta, dpi=300, units=c("mm"), width=300, height=300)
```


## Figure S5

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM25
#| fig-cap: "Figure S5. Comparison of log-transformed leaf mass per area (LMA) by species and elevation zone. A) Compares logLMA means by species and elevation zone: LOW (<1220 m), MID (1221 - 1828 m), HIGH (> 1829 m). B) Change in logLMA per species as elevation increases. The black trend line represents the fitted model for all data points, see main text for *R^2^~adj~* and *p* values. Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < 0.01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

########## Labels ###########
#Legend title label
legend_title <- "Species"
# Species labels
labels_list <- c("M. laciniatus", "M. nasutus", "M. guttatus")
names(labels_list) <- c("L", "N", "G")
#############################

#ggpubr package used for comparisons tests
lma_compare_means <- compare_means(
  logLMA ~ Species,
  data = leaf_traits_noB,
  group.by = "Elevation_cat",
  method = "wilcox.test",
  p.adjust.method = "BH",
  paired = FALSE,
  alternative = "two.sided",
  mu = 0,
  var.equal = FALSE,
  conf.level = 0.95
)

#lma_compare_means
#Report with `compare_means` because it is easier to plot with `ggpubr` package.

#Error bars for plots
errbar_lims <- group_by(leaf_traits_noB, Site) |>
  summarize(
    mean = mean(logLMA),
    se = sd(logLMA) / sqrt(n()),
    upper = mean + (2 * se),
    lower = mean - (2 * se)
  )
#Comparisons
species_comparisons <- list(
  c("M. nasutus", "M. laciniatus"),
  c("M. nasutus", "M. guttatus"),
  c("M. laciniatus", "M. guttatus")
)

#By Site discrete comparison
lma <- ggplot(leaf_traits_noB, aes(y = logLMA, x = Species)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Species), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    labels = c('M. laciniatus', 'M. nasutus', 'M. guttatus'),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  #scale_color_manual(aes(color = "black")) +
  theme_lfspp_discrete +
  labs(y = "log10[Leaf Mass per Area (LMA) (mg/mm)]", x = "") +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    #label.y = c(0.4, 0.6, 1),
    label = "p.signif"
  ) +
  facet_wrap(~Elevation_cat)

#lma
#ggsave(filename = file.path(path, "figures/lma_violin.png"), plot = lma, dpi=300, units=c("mm"), width=250, height=150)

# By Elevation continuous comparison
# Linear model summary
# summary(lm(logLMA ~ Elevation_m, data = leaf_traits_noB |>
#              filter(!logLMA == "-Inf"), na.action = na.omit)) # Remove -Inf values
elev_lma_p <- format.p(
  cor.test(
    leaf_traits_noB$logLMA,
    leaf_traits_noB$Elevation_m,
    na.action = na.omit
  )$p.value
) # This is just to make the p-value nicer.

# logLMA by Elevation
lma_cont <- ggplot(
  leaf_traits_noB |> # Removing infinite values
    filter(!logLMA == "-Inf"),
  aes(y = logLMA, x = Elevation_m)
) +
  geom_jitter(aes(color = Species), alpha = 0.6, size = 2.5) +
  geom_smooth(aes(color = Species), method = lm, se = FALSE, na.rm = F) +
  geom_smooth(
    method = lm,
    se = TRUE,
    level = 0.95,
    na.rm = F,
    color = "black"
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    parse = TRUE,
    label.x.npc = "left",
    vstep = 0.08
  ) +
  scale_color_manual(
    labels = c('M. laciniatus', 'M. nasutus', 'M. guttatus'),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  theme_lfspp_continuous +
  labs(y = "log10[Leaf Mass per Area (LMA)]", x = "Elevation (m.a.s.l.)")

#lma_cont
#ggsave(filename = file.path(path, "figures/lma_elev.png"), plot = lma_cont, dpi=300, units=c("mm"), width=300, height=150)

# Joining plots
lma_elevation <- ggarrange(
  lma_cont,
  lma,
  ncol = 1,
  nrow = 2,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A", "B"),
  heights = c(1, 1)
)

lma_elevation

ggsave(
  filename = file.path(path, "figures/supp_fig5.jpg"),
  plot = lma_elevation,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 250
)
```

## Figure S6

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 5
#| label: suppfigMIM26
#| fig-cap: "Figure S6. Change in log-transformed anthocyanin content index (ACI) by species and elevation zone. A) Compares logACI means by species and elevation zone: LOW (<1220 m), MID (1221 - 1828 m), HIGH (> 1829 m). The black trend line represents the fitted model for all data points, see main text for *R^2^~adj~* and *p* values. B) Change in logLPS per species as elevation increases.  Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < .01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

#ggpubr package used for comparisons tests
aci_compare_means <- compare_means(
  logACI ~ Species,
  data = leaf_traits_noB,
  group.by = "Elevation_cat",
  method = "wilcox.test",
  p.adjust.method = "BH",
  paired = FALSE,
  alternative = "two.sided",
  mu = 0,
  var.equal = FALSE,
  conf.level = 0.95
)

#aci_compare_means
#Roughly the same outcome as pairwise_t_test.

#Report with `compare_means` because it is easier to plot with `ggpubr` package.
errbar_lims <- group_by(leaf_traits_noB, Species) |>
  summarize(
    mean = mean(logACI),
    se = sd(logACI) / sqrt(n()),
    upper = mean + (2 * se),
    lower = mean - (2 * se)
  )

#Comparisons
species_comparisons <- list(
  c("M. nasutus", "M. laciniatus"),
  c("M. nasutus", "M. guttatus"),
  c("M. laciniatus", "M. guttatus")
)


aci <- ggplot(leaf_traits_noB, aes(y = logACI, x = Species)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Species), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = -1)
  ) +
  theme_lfspp_discrete +
  theme(axis.text.y = element_text(size = 10)) +
  labs(y = paste("log[Anthocyanin Content Index (ACI)]"), x = "") +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    #label.y = c(0.4, 0.6, 1),
    label = "p.signif"
  ) +
  facet_wrap(~Elevation_cat)

#aci
#ggsave(filename = file.path(path, "figures/aci_violin.png"), plot = aci, dpi=300, units=c("mm"), width=300, height=150)

# By Elevation continuous comparison
# Linear model summary
#summary(lm(logACI~ Elevation_m, data = leaf_traits_noB, na.action = na.omit))
elev_aci_p <- format.p(
  cor.test(
    leaf_traits_noB$logACI,
    leaf_traits_noB$Elevation_m,
    na.action = na.omit
  )$p.value
) # This is just to make the p-value nicer.

# logACI by Elevation
aci_cont <- ggplot(leaf_traits_noB, aes(y = logACI, x = Elevation_m)) +
  geom_jitter(aes(color = Species), alpha = 0.6, size = 2.5) +
  geom_smooth(aes(color = Species), method = lm, se = FALSE, na.rm = F) +
  geom_smooth(
    method = lm,
    se = TRUE,
    level = 0.95,
    na.rm = F,
    color = "black"
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    parse = TRUE,
    label.x.npc = "left",
    vstep = 0.08
  ) +
  scale_color_manual(
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  theme_lfspp_continuous +
  theme(
    #plot.caption = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  ) +
  labs(
    size = 10,
    y = paste("log[Anthocyanin Content Index (ACI)]"),
    x = "Elevation (m.a.s.l.)"
    # caption = (expression(
    #   paste("ACI = % ", italic("A") [lambda], " at 931 nm / % ", italic("A") [lambda], " at 530 nm")
    # ))
  ) +
  scale_y_continuous(expand = c(0.07, 0.07))

#aci_cont
#ggsave(filename = file.path(path, "figures/aci_cont.png"), plot = aci_cont, dpi=300, units=c("mm"), width=300, height=150)

# Joining plots
aci_elevation <- ggarrange(
  "",
  aci_cont,
  "",
  aci,
  ncol = 1,
  nrow = 4,
  heights = c(0.1, 1.1, 0.1, 1.1),
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A", "", "B")
)

aci_elevation
ggsave(
  filename = file.path(path, "figures/supp_fig6.jpg"),
  plot = aci_elevation,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 250
)
```

## Figure S7

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM27
#| fig-cap: "Figure S7. Change in log-transformed leaf lobe index (LBI) by species and elevation zone. A) Compares logLBI means by species and elevation zone: LOW (<1220 m), MID (1221 - 1828 m), HIGH (> 1829 m). The black trend line represents the fitted model for all data points, see main text for *R^2^~adj~* and *p* values. B) Change in logLPS per species as elevation increases. Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < .01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

#ggpubr package used for comparisons tests
lbi_compare_means <- compare_means(
  logLBI ~ Species,
  data = leaf_traits_noB,
  group.by = "Elevation_cat",
  method = "wilcox.test",
  p.adjust.method = "BH",
  paired = FALSE,
  alternative = "two.sided",
  mu = 0,
  var.equal = FALSE,
  conf.level = 0.95
)

#lbi_compare_means

#Roughly the same outcome as pairwise_t_test.

#Report with `compare_means` because it is easier to plot with `ggpubr` package.
errbar_lims <- group_by(leaf_traits_noB, Species) |>
  summarize(
    mean = mean(logLBI),
    se = sd(logLBI) / sqrt(n()),
    upper = mean + (2 * se),
    lower = mean - (2 * se)
  )

#Comparisons
species_comparisons <- list(
  c("M. nasutus", "M. laciniatus"),
  c("M. nasutus", "M. guttatus"),
  c("M. laciniatus", "M. guttatus")
)

lbi <- ggplot(leaf_traits_noB, aes(y = logLBI, x = Species)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Species), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    labels = c('M. laciniatus', 'M. nasutus', 'M. guttatus'),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  theme_lfspp_discrete +
  labs(y = "log[Leaf Lobe Index (LBI)", x = "") +
  #scale_x_discrete(labels = c('M. laciniatus', 'M. nasutus', 'M. guttatus')) +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  facet_wrap(~Elevation_cat)

#lbi
#ggsave(filename = file.path(path, "figures/lbi_violin.png"), plot = lbi, dpi=300, units=c("mm"), width=300, height=150)

# By Elevation continuous comparison
# Linear model summary
#summary(lm(logLBI~ Elevation_m, data = leaf_traits_noB, na.action = na.omit))
elev_lbi_p <- format.p(
  cor.test(
    leaf_traits_noB$logLBI,
    leaf_traits_noB$Elevation_m,
    na.action = na.omit
  )$p.value
) # This is just to make the p-value nicer.

# logLBI by Elevation
lbi_cont <- ggplot(leaf_traits_noB, aes(y = logLBI, x = Elevation_m)) +
  geom_jitter(aes(color = Species), alpha = 0.6, size = 2.5) +
  geom_smooth(aes(color = Species), method = lm, se = FALSE, na.rm = F) +
  geom_smooth(
    method = lm,
    se = TRUE,
    level = 0.95,
    na.rm = F,
    color = "black"
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    rr.digits = 2,
    p.digits = 3,
    parse = TRUE,
    label.x.npc = "left",
    vstep = 0.08
  ) +
  scale_color_manual(
    labels = c('*M. laciniatus*', '*M. nasutus*', '*M. guttatus*'),
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  theme_classic(base_size = 16) +
  theme(
    legend.title = element_text(face = "bold"),
    panel.border = element_rect(linetype = "blank", fill = NA),
    legend.text = element_markdown(size = 14), #Make legend text italic
    axis.text.x = element_text(size = 14),
    legend.position = "bottom"
  ) +
  labs(y = "log[Leaf Lobe Index (LBI)]", x = "Elevation (m.a.s.l.)") +
  scale_y_continuous(expand = c(0.07, 0.07))

#lbi_cont
#ggsave(filename = file.path(path, "figures/lbi_elev.png"), plot = lbi_cont, dpi=300, units=c("mm"), width=300, height=150)

# Joining plots
lbi_elevation <- ggarrange(
  lbi_cont,
  lbi,
  ncol = 1,
  nrow = 2,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A", "B"),
  heights = c(1, 1)
)

lbi_elevation

ggsave(
  filename = file.path(path, "figures/supp_fig7.jpg"),
  plot = lbi_elevation,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 250
)
```

## Figure S8

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM28
#| fig-cap: "Figure S8. Change in log-transformed leaf-punch strength (LPS), a measure of leaf toughness, by species and elevation zone. A) Compares logLPS means by species and elevation zone: LOW (<1220 m), MID (1221 - 1828 m), HIGH (> 1829 m). The black trend line represents the fitted model for all data points, see main text for *R^2^~adj~* and *p* values. B) Change in logLPS per species as elevation increases. Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < .01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

#ggpubr package used for comparisons tests
lps_compare_means <- compare_means(
  logLPS ~ Species,
  data = leaf_traits_noB,
  group.by = "Elevation_cat",
  method = "wilcox.test",
  p.adjust.method = "BH",
  paired = FALSE,
  alternative = "two.sided",
  mu = 0,
  var.equal = FALSE,
  conf.level = 0.95
)

#lps_compare_means

#Report with `compare_means` because it is easier to plot with `ggpubr` package.
errbar_lims <- group_by(leaf_traits_noB, Species) |>
  summarize(
    mean = mean(logLPS),
    se = sd(logLPS) / sqrt(n()),
    upper = mean + (2 * se),
    lower = mean - (2 * se)
  )


lps <- ggplot(leaf_traits_noB, aes(y = logLPS, x = Species)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Species), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    values = met.brewer(
      "Isfahan2",
      n = 4,
      type = "discrete",
      direction = c(-1)
    )
  ) +
  theme_lfspp_discrete +
  labs(
    y = expression(paste("Leaf punch strength (LPS)(N mm", "^-1)")),
    x = ""
  ) +
  #scale_x_discrete(labels = c('M. laciniatus', 'M. nasutus', 'M. guttatus')) +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    #label.y = c(0.4, 0.6, 1),
    label = "p.signif"
  ) +
  facet_wrap(~Elevation_cat)

#lps
#ggsave(filename = file.path(path, "figures/lps_violin.png"), plot = lps, dpi=300, units=c("mm"), width=300, height=150)

# By Elevation continuous comparison
# Linear model summary
#summary(lm(logLPS ~ Elevation_m, data = leaf_traits_noB, na.action = na.omit))
elev_lps_p <- format.p(
  cor.test(
    leaf_traits_noB$logLPS,
    leaf_traits_noB$Elevation_m,
    na.action = na.omit
  )$p.value
) # This is just to make the p-value nicer.

# logLPS by Elevation
lps_cont <-
  ggplot(leaf_traits_noB, aes(y = logLPS, x = Elevation_m)) +
  geom_jitter(aes(color = Species), alpha = 0.6, size = 2.5) +
  geom_smooth(aes(color = Species), method = lm, se = FALSE, na.rm = F) +
  geom_smooth(
    method = lm,
    se = TRUE,
    level = 0.95,
    na.rm = F,
    color = "black"
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    rr.digits = 2,
    p.digits = 3,
    parse = TRUE,
    label.x.npc = "left",
    vstep = 0.08
  ) +
  scale_color_manual(
    values = met.brewer(
      "Isfahan2",
      n = 4,
      type = "discrete",
      direction = c(-1)
    )
  ) +
  theme_lfspp_continuous +
  labs(
    y = expression(paste("Leaf punch strength (LPS) (N mm", "^-1)")),
    x = "Elevation (m.a.s.l.)"
  ) +
  scale_y_continuous(expand = c(0.07, 0.07))

#lps_cont
#ggsave(filename = file.path(path, "figures/lps_elev.png"), plot = lps_cont, dpi=300, units=c("mm"), width=300, height=150)

# Joining plots
lps_elevation <- ggarrange(
  "",
  lps_cont,
  "",
  lps,
  ncol = 1,
  nrow = 4,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A", "", "B"),
  heights = c(0.1, 1, 0.1, 1)
)

lps_elevation

ggsave(
  filename = file.path(path, "figures/supp_fig8.jpg"),
  plot = lps_elevation,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 250
)
```

## Figure S9

```{r}
#| echo: false
#| eval: true
#| tidy: true
#| fig-width: 6
#| fig-height: 6
#| label: suppfigMIM29
#| fig-cap: "Figure S9. Change in log-transformed leaf thickness (LT) (𝜇m) by species and elevation zone. A) Compares logLT means by species and elevation zone: LOW (<1220 m), MID (1221 - 1828 m), HIGH (> 1829 m). The black trend line represents the fitted model for all data points, see main text for *R^2^~adj~* and *p* values. B) Change in logLT per species as elevation increases. Significance levels are represented by *ns* (not significant) and asterisks [*p* < .05 (\\*), *p* < .01 (\\*\\*), *p* < .001 (\\*\\*\\*), and *p* < .0001 (\\*\\*\\*\\*)]."

#ggpubr package used for comparisons tests
lt_compare_means <- compare_means(
  logLT ~ Species,
  data = leaf_traits_noB,
  group.by = "Elevation_cat",
  method = "wilcox.test",
  p.adjust.method = "BH",
  paired = FALSE,
  alternative = "two.sided",
  mu = 0,
  var.equal = FALSE,
  conf.level = 0.95
)

#lt_compare_means
#Roughly the same outcome as pairwise_t_test.

#Report with `compare_means` because it is easier to plot with `ggpubr` package.
errbar_lims <- group_by(leaf_traits_noB, Species) |>
  summarize(
    mean = mean(logLT),
    se = sd(logLT) / sqrt(n()),
    upper = mean + (2 * se),
    lower = mean - (2 * se)
  )

#Comparisons
species_comparisons <- list(
  c("M. nasutus", "M. laciniatus"),
  c("M. nasutus", "M. guttatus"),
  c("M. laciniatus", "M. guttatus")
)

lt <- ggplot(leaf_traits_noB, aes(y = logLT, x = Species)) +
  geom_point(
    aes(color = Species),
    alpha = 0.2,
    size = 1.7,
    position = position_jitter(w = 0.2)
  ) +
  geom_boxplot(aes(color = Species), alpha = 0.5, outlier.alpha = 0.0) +
  scale_color_manual(
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  theme_lfspp_discrete +
  labs(y = expression(paste("log[Leaf thickness (LT) (", mu, "m)]")), x = "") +
  #scale_x_discrete(labels = c('M. laciniatus', 'M. nasutus', 'M. guttatus')) +
  stat_compare_means(
    comparisons = species_comparisons,
    method = "wilcox.test",
    #label.y = c(0.4, 0.6, 1),
    label = "p.signif",
    hide.ns = TRUE
  ) +
  facet_wrap(~Elevation_cat)

#lt
#ggsave(filename = file.path(path, "figures/lt_violin.png"), plot = lt, dpi=300, units=c("mm"), width=300, height=150)

# By Elevation continuous comparison
# Linear model summary
#summary(lm(logLT ~ Elevation_m, data = leaf_traits_noB, na.action = na.omit))
#elev_lt_p <- format.p(cor.test(leaf_traits_noB$logLT, leaf_traits_noB$Elevation_m, na.action = na.omit)$p.value) # This is just to make the p-value nicer.

# logLT by Elevation
lt_cont <- ggplot(leaf_traits_noB, aes(y = logLT, x = Elevation_m)) +
  geom_jitter(aes(color = Species), alpha = 0.6, size = 2.5) +
  geom_smooth(aes(color = Species), method = lm, se = FALSE, na.rm = F) +
  geom_smooth(
    method = lm,
    se = TRUE,
    level = 0.95,
    na.rm = F,
    color = "black"
  ) +
  stat_poly_eq(
    formula = y ~ x,
    aes(
      color = Species,
      label = paste(
        #..eq.label..,
        ..adj.rr.label..,
        #"~italic(p) ==",
        ..p.value.label..,
        #signif(..p.value.., digits = 4),
        sep = "*`,`~~"
      )
    ),
    rr.digits = 2,
    p.digits = 3,

    parse = TRUE,
    label.x.npc = "left",
    vstep = 0.08
  ) +
  scale_color_manual(
    values = met.brewer("Isfahan2", n = 4, type = "discrete", direction = c(-1))
  ) +
  theme_lfspp_continuous +
  labs(
    y = expression(paste("log[Leaf thickness (LT) (", mu, "m)]")),
    x = "Elevation (m.a.s.l.)"
  ) +
  scale_y_continuous(expand = c(0.07, 0.07))

#lt_cont
#ggsave(filename = file.path(path, "figures/lps_elev.png"), plot = lps_cont, dpi=300, units=c("mm"), width=300, height=150)

# Joining plots
lt_elevation <- ggarrange(
  lt_cont,
  lt,
  ncol = 1,
  nrow = 2,
  common.legend = TRUE,
  legend = "bottom",
  labels = c("A", "B"),
  heights = c(1, 1)
)
lt_elevation
ggsave(
  filename = file.path(path, "figures/supp_fig9.jpg"),
  plot = lt_elevation,
  dpi = 300,
  units = c("mm"),
  width = 250,
  height = 250
)
```


## Table S1

::: {#supptbl-1}
```{r}
#| label: supp-table-s1-code
#| echo: false

# Sites and Species sampled
table_data <- leaf_traits_noB |>
  group_by(Site, Species, Longitude, Latitude) |>
  summarise(n = n_distinct(Unique_ID), .groups = "drop") |>
  pivot_wider(names_from = Species, values_from = n) |>
  left_join(
    extra_site_details |>
      select(Unique_ID, Site_name, Forest_Park) |>
      distinct(),
    by = c("Site" = "Unique_ID")
  ) |>
  filter(!Site == "FVWB" & !Site == "LYFB") |> # Removing out of bounds

  select(
    Site,
    Site_name,
    Forest_Park,
    Latitude,
    Longitude,
    `M. laciniatus`,
    `M. nasutus`,
    `M. guttatus`
  ) |>
  mutate(
    Forest_Park = str_replace_all(Forest_Park, "_", " ")
  ) |>
  rename(
    "Site Name" = Site_name,
    "Forest/Park" = Forest_Park
  )

totals <- c(
  sum(table_data$`M. laciniatus`, na.rm = TRUE),
  sum(table_data$`M. nasutus`, na.rm = TRUE),
  sum(table_data$`M. guttatus`, na.rm = TRUE)
)

tableS1 <- table_data |>
  huxtable::huxtable() |>
  huxtable::set_number_format(everywhere, 4:5, value = "%5.6f") |>
  huxtable::set_italic(1, 6:8, TRUE) |>
  set_all_padding(10) |>
  set_outer_padding(0) |>
  set_width(1.1) |>
  theme_article() |>
  huxtable::set_caption(
    "Table S1: Number of individuals collected per population/site and species."
  ) |>
  huxtable::as_flextable(column_to_header = TRUE) |>
  flextable::add_footer_lines("") |>
  flextable::add_footer_lines(
    values = as_paragraph(
      "Total | | ",
      as_chunk(paste(totals, collapse = " | "))
    )
  )

tableS1
# knitr::kable(
#   caption = "Table S1. Number of individuals collected per population/site and species."
# )
```

Number of individuals collected per population/site and species.
:::


## Table S2

::: {#supptbl-2}
```{r}
#| label: supp-table-s2-code
tableS2 <- phyloseq_ntaxa_by_tax(
  ps_clean_3,
  TaxRank = "Phylum",
  relative = F,
  add_meta_data = T
) |>
  as.data.frame() |>
  mutate(Phylum = str_remove(Phylum, "^.*?(_|-).")) |>
  group_by(Species, Phylum) |>
  summarise(Occurrence = n(), .groups = "drop") |>
  gt(rowname_col = "Phylum", groupname_col = "Species") |>
  fmt_number(columns = Occurrence, decimals = 0) |>
  tab_options(
    table.font.names = "Tex Gyre Termes",
    heading.title.font.size = "medium",
    row_group.as_column = TRUE
  ) |>
  tab_stub_indent(
    rows = everything(),
    indent = 3
  ) |>
  # tab_header(
  #   title = "Table S2. Prevalence of phyla in samples per species."
  # ) |>
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_row_groups()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_footnote(
    footnote = "Occurrence represents the number of samples where each phylum was detected per species."
  )

tableS2
```

Prevalence of phyla in samples per species.
:::

## Table S3

::: {#supptbl-3}
```{r}
#| label: supp-table-s3-code
#| echo: false
#| eval: true
#| tidy: true

#dbRDA table  (results from terms are the same as adonis2 but without R2)
anova_m5hell_term <- readRDS(
  "data/clean_data/statistics/anova_m5hell_term.rds"
)

anova_table3 <- tidy(anova_m5hell_term) |>
  as.data.frame() |>
  mutate("p-value" = str_replace_all(p.value, "0.001", "***")) |>
  rename(
    Term = term,
    "Sum. of Sqs." = SumOfSqs,
    "F-value" = statistic,
    "Df" = df
  ) |>
  select(Term, Df, "Sum. of Sqs.", "F-value", "p-value")

# Table

tableS3 <- flextable(anova_table3) |>
  flextable::set_table_properties(width = .5, layout = "autofit") |>
  flextable::colformat_double(digits = 3) |>
  flextable::style(i = c(1:4), j = 1, pr_t = fp_text(bold = FALSE)) |>
  flextable::bold(
    i = c(1),
    j = c(1:5),
    bold = TRUE,
    part = "header"
  ) |>
  flextable::italic(
    i = c(5),
    j = 1,
    italic = TRUE,
    part = "body"
  ) |>
  flextable::font(
    i = NULL,
    j = NULL,
    fontname = "Tex Gyre Termes",
    part = "all",
    cs.family = fontname,
    hansi.family = fontname,
    eastasia.family = fontname
  ) |>
  flextable::set_caption(
    "Table S3. Permutational ANOVA for the best-fit dbRDA model with hybrid cross-types."
  ) |>
  flextable::border(
    i = 5,
    j = c(1:5),
    border.top = fp_border(color = "black", width = 1)
  ) |>
  flextable::add_footer_lines(
    value = as_paragraph(
      "Significance levels are represented by ns (not significant) and asterisks [p < .05 (*), p < .01 (**), p < .001 (***), and p < .0001 (****)]."
    )
  )

tableS3
```

Permutational ANOVA for the best-fit dbRDA model with hybrid cross-types. Significance levels are represented by asterisks [*p* < .05 (\*), *p* < .01 (\*\*), *p* < .001 (\*\*\*), and *p* < .0001 (\*\*\*\*)].
:::

## Supplementary methods

### Leaf trait measurements: instruments and protocol detail

We cleaned plants with tap water to remove all soil and debris remnants from the leaves and roots. We removed all healthy leaves (~ 5 - 10) from the stems and took three measurements per trait from three haphazardly selected leaves from individual plants, except for LBI, where we measured only one leaf per plant. To calculate LBI we used a transparency film to hold the leaf in place and flatten, after which we took a digital photograph and then followed Ferris et al. [-@ferris2015]. We measured ACI content with ACM-200plus (Opti-Sciences Inc. Hudson, New Hampshire, U.S.A.) on haphazardly selected locations of the leaf surface (working from the petiole out to the leaf tip) [@tellez2022]. The ACM-200 calculates an ACI value from the ratio of % transmittance at 931 nm/% transmittance at 525 nm [@opti-sciencesinc], effectively accounting for leaf thickness. We measured LT (μm) with a Mitutoyo 7327 Micrometer Gauge (Mitutoyo, Takatsu-ku, Kawasaki, Japan) on haphazardly selected locations of the leaf lamina, taking care to avoid major and secondary veins. We used an Imada DST-11a digital force gauge (Imada Inc., Northbrook, IL, United States) to measure LPS on the lamina of each leaf selected, avoiding minor leaf veins when possible [@tellez2022]. It functions by conducting punch-and-die tests with a sharp-edged cylindrical steel punch (2.0 mm diameter) and a steel die with a sharp-edged aperture of small clearance (0.05 mm). Once LPS was measured, we used a 4 mm diameter punch hole to puncture disks for LMA measurements, collecting one disk per leaf. The dried disk punches were shipped to Tulane University, New Orleans, LA, USA to dry at 60℃ for 48-72 hours before being weighed.

### Tissue preservation: full protocol

Upon completion of the leaf trait measurements, we prepared and preserved samples at the UC Merced Yosemite Field Station. We started by removing the main vein and margins from photosynthetic tissue. The leaf lamina was haphazardly cut with a sterile blade into 2 mm wide strips in parallel to the main vein [@arnold2003; @tellez2022; @higgins2014]. Leaf strips were then sterilized with sequential washes in 95% EtOH (10 s), 0.5% sodium hypochlorite (NaOCl) (60 s), and 70% EtOH (60 s) and air dried under sterile conditions. Due to the small size of monkeyflower plants, the maximum amount of leaf lamina was preserved in sterile 15 mL tubes with ~ 10 mL CTAB solution (1 M Tris-HCl pH 8, 5 M NaCl, 0.5 M EDTA, and 20 g CTAB). All leaf tissue handling was performed in a sterile environment with an alcohol burner lamp inside a portable biosafety cabinet. All surfaces were previously sterilized sequentially with 0.5% NaOCl, 95% EtOH, and 70% EtOH. We surface sterilized surfaces and instruments in between sample handling to prevent cross contamination.

### Amplicon sequencing: reagents, instruments, and reaction detail

We stored leaf tissue in CTAB solution for 2 months at room temperature before extracting DNA at Tulane University. To prepare for the DNA extraction procedure, we decontaminated all instruments, materials, and surfaces in the biosafety cabinet with 0.5% NaOCl, 70% EtOH, and 95% EtOH, and subsequently treated them with UV light for 30 minutes. We subsampled 0.2 - 0.3 g of leaf tissue from each sample and placed it into sterile 2 mL tubes containing an assortment of beads: 3.2 mm stainless steel beads (Next Advance, Cat# SSB32), 100 µL stainless steel bead blend, 0.9-2.0 mm (NextAdvance, Cat# SSB14B) and 2-3 of the autoclaved 2 mm zirconium oxide beads (Next Advance, Cat# ZRoB20). We then lyophilized samples for 72 hours to fully remove CTAB content from tissue, submerged the sample tubes in liquid nitrogen for 30 s, and homogenized samples at 30 Hz for 3 minutes in a TissueLyser LT (QIAGEN, Valencia, CA, USA). We stored samples at −20℃ until DNA extraction.

We used a DNA extraction protocol for high-molecular-weight DNA extraction adapted from Russo et al. [-@russo2022]. Briefly, it is a CTAB:chloroform:isoamyl extraction combined with a solid-phase reversible immobilization (SPRI) bead step [@russo2022; @rohland2012; @liu2023]. Protocol modifications allowed us to optimize extractions for fungal DNA from preserved leaf tissue [see details in @aponterolon2023a]. After all genomic DNA was extracted, we quantified the DNA using a Quant-iT® dsDNA HS Assay kit with a Qubit Flourometer (Thermo Fisher Scientific, Waltham, MA, USA, Cat# Q33120) and followed a two-step amplification approach described by Sarmiento et al. [-@sarmiento2017] and U'Ren & Arnold [-@uren2017]. Every sample was amplified in three parallel reactions at annealing temperatures of 52℃, 54℃, and 56℃ to amplify a wide range of fungal taxa and reduce amplification bias for short ITS sequences [@uren2017; @lumibao2018]. Each PCR (PCR1) reaction contained 2 µL of sample DNA template. We visualized PCR1 reactions with SYBR™ Safe DNA Gel Stain (Thermo Fisher Scientific, Waltham, MA, USA, Cat# S33102) on a 2% agarose gel [@oita2021]. We combined 5 µL of amplicon product from parallel reactions into a single tube per sample and purified using Sera-Mag™ SpeedBead Carboxylate-Modified Magnetic Particles (Hydrophobic) (Thermo Fisher Scientific, Waltham, MA, USA, Cat#09-981-123) prepared as per [-@rohland2012 and -@liu2023] and used a ratio of 1.2x:1 with 80% EtOH following manufacturer's instructions. We used 3 µL of PCR1 product from samples, DNA extraction controls, and PCR1 negative controls for a second PCR (PCR2) with barcoded adapters (IDT, Coralville, Iowa, USA). Each PCR2 reaction (total 30 µL) contained 1X Phusion Flash High Fidelity PCR Master Mix (Thermo Fisher Scientific, Waltham, MA, USA, Cat# F548L), 0.075 µM of barcoded primers (forward and reverse pooled at an initial concentration of 2 µM), and 0.20 mg/mL of BSA (Thermo Fisher Scientific, Waltham, MA, USA, Cat# B14), following [-@sarmiento2017 and -@uren2017]. Before final pooling for sequencing, we purified and concentrated amplicons using SPRI beads to a total volume of 20 µL. We quantified PCR2 product with a Quant-iT™ PicoGreen™ dsDNA Assay Kit (Thermo Fisher Scientific, Waltham, MA, USA, Cat# P7589) with a BioTek Synergy LX plate reader (Agilent, Santa Clara, CA) and combined equimolar amounts of libraries, including DNA extraction controls, PCR1, and PCR2 negative controls, into a 10 nM library pool. We did not detect any contamination visually or fluorometrically. Throughout all these steps, we used a separate set of sterile pipettes, tips, and equipment to reduce contamination in a designated PCR area to restrict contact with pre-PCR materials [@oita2021].

### Illumina TruSeq adapters

The modified primers for the first PCR (adapter ligation and ITS1 amplification) were as follows: 5’-CAC TCT TTC CCT ACA CGA CGC TCT TCC GAT CTC TTG GTC ATT TAG AGG AAG TAA-3’ (forward) and 5’-GTG ACT GGA GTT CAG ACG TGT GCT CTT CCG ATC TGC TGC GTT CTT CAT CGA TGC-3’ (reverse).
See [@gardes1993] and [@white1990] for more details.

### Supplementary Literature Cited

Gardes, M., & Bruns, T. D. (1993). ITS primers with enhanced specificity for basidiomycetes ‐ application to the identification of mycorrhizae and rusts. Molecular Ecology, 2(2), 113–118. https://doi.org/10.1111/j.1365-294X.1993.tb00005.x

White, T. J., Bruns, T.D., Lee, S. B., & Taylor, J. W. (1990). Amplification and direct sequencing of fungal ribosomal RNA genes for phylogenetics (pp. 315–322).
